---
title: "MAR Analysis Workflow"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{MAR Analysis Workflow}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown:
        wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    fig.align = "center",
    fig.width = 6,
    fig.height = 4.5,
    dpi = 150,
    out.width = "100%"
)
rm(list = ls())
```

# The standard `mar` workflow

## Overview

This vignette walks through the standard workflow for a mutations-area relationship (MAR)
and genetic-diversity-area relationship (GDAR) analysis in the `mar` package, using the
bundled *Arabidopsis thaliana* 1001 Genomes data. The workflow comprises five modules:
read genotype and coordinate files into the core `genomaps` object; compute the site
frequency spectrum and predict the MAR/GDAR theoretically; build it empirically by
sampling the geographic range; simulate the extinction process; and fit the power-law
model $M = cA^z$ to estimate the scaling factor $z$, which describes the rate of genetic
diversity loss.

For the sampling theory behind the relationship, and for guidance on which $z$ to carry
into a conservation prediction, see `vignette('mar-theory')`.

```{r setup}
library(mar)
```


## Step 0: Load example data

The `mar` package provides an example dataset in the `genomaps` data format ready for analysis.

```{r example}
data(gm1001g)
print(gm1001g)

# gm1001g is a list of two objects, to obtain marmaps and margeno objects
gm1001g$geno
gm1001g$maps

# Tips: all operations in the `mar` package take `genomaps` as the input.
# Therefore, you rarely need to extract the marmaps and margeno objects yourself.
```

If you do not need to load your own data, skip to [Step 3: Explore the Site Frequency Spectrum (SFS)](#s3).

Otherwise, continue to [Step 1: Import raw data](#s1)

## Step 1: Import raw data {#s1}

### 1a: Import genotype data using `text_parser()`

`text_parser()` reads the genotype matrix and returns a `margeno` object
containing the genotype data alongside sample and variant metadata.

The genotype file is a delimited text file of alternative allele counts, one row per
variant and one column per sample, with no header and no row names. Entries are integers
from 0 to `ploidy`, so 0, 1 or 2 under the diploid default, and `NA` marks a missing
genotype. To avoid violating the model assumptions, sites must be biallelic SNPs or
invariant sites. Two optional files supply the metadata: a single-column file of sample
identifiers in the same order as the genotype columns, and a two-column `CHR`/`POS` file
of variant positions.

```{r import genotype}
# Tips: Begin by locating and importing the bundled example data included with the
# `mar` package. `system.file()` resolves the path to files stored in `inst/extdata/`,
# ensuring the vignette works regardless of where the package is installed.

# Locate sample genotypes
# Tips: text_parser reads both gzipped and uncompressed text files.
genofile <- system.file("extdata", "1001g_genotypes.txt.gz", package = "mar")
# Optional: text_parser also reads sample names and chromosome positions if provided
samplefile <- system.file("extdata", "1001g_accessions.txt", package = "mar")
chrposfile <- system.file("extdata", "1001g_chrpos.txt", package = "mar")
# Tips: for your own data, change the variables above to the corresponding filepaths

# Read genotype
# Tips: for vcf parsing, see Alternative 1: Reading VCF files
genodata <- text_parser(
    geno.fn = genofile,
    samp.fn = samplefile,
    pos.fn = chrposfile
)

print(genodata)
```

**Alternative workflow**:
    `vcf_parser()` handles reading of VCF files, see [Alternative 1: Reading VCF files](#a1).

### 1b: Import geographic data using `lonlat_parser()` {#s1b}

`lonlat_parser()` reads the coordinate file (TXT, CSV, or TSV) and returns a `marmaps` object.

The coordinate file holds exactly three columns, `ID`, `LON` and `LAT` (or `LONGITUDE` and
`LATITUDE`), with a header and one row for every sample in the genotype matrix. Alongside
the coordinates, the `marmaps` object stores a raster map that summarises where samples
fall and serves as the reference grid for every downstream analysis. Resolution can be set
manually or, as here, determined automatically from the map extent and sample density.

```{r import coordinates}
# Locate sample coordinates
lonlatfile <- system.file("extdata", "1001g_lonlat.txt", package = "mar")
# Read coordinates
# Tips: lonlat_parser will assume the coordinates are longitude and latitude following ESPG: 4326 (WGS84) by default
# and automatically determine the map resolution following equation 12 in (Hengl 2006)
mapsdata <- lonlat_parser(lonlatfile, mapcrs = "EPSG:8857")

print(mapsdata)
# Optional: to inspect the reference map created
plot(mapsdata)
```

**Alternative workflow**:
    To change map resolution and projection, see [Alternative 2: Customizing projection and map resolution](#a2).

## Step 2: Construct `genomaps` object using `genomaps()`

After successful importing of genotypes and geographic data, it is now time to construct the main object for `mar` workflow.

`genomaps()` combines the `margeno` and `marmaps` objects into a single analysis
object used by all downstream functions.

```{r genomaps}
gm1001g_load <- genomaps(genodata, mapsdata)
# The created gm1001g_load is the same as the provided example gm1001g
all.equal(gm1001g_load, gm1001g)
rm(genodata, mapsdata, gm1001g_load)
```


## Step 3: Site-frequency spectrum computation and theoretical MAR/GDAR prediction {#s3}

### 3a: Build SFS using `sfs()`

The site frequency spectrum is a histogram of mutation abundances and the basic summary of
a genetic dataset. In a `genomaps` object of $J$ individuals with ploidy $x$ there are $xJ$
chromosomes and $L$ genotyped biallelic sites. At any one site, count how many of the $xJ$
chromosomes carry the mutation and call that count $k$: `sfs()` reports how many sites have
$k = 0$ (invariants), how many have $k = 1$ (singletons), and so on. Most mutations sit at
low frequency and few reach high frequency, and this commonness of rarity is what
ultimately generates the MAR.

The observed SFS summarises how allele counts are distributed across sites, and `expsfs()`
returns the neutral coalescent expectation $E(k) = \theta/k$ as a baseline for comparison.

```{r sfs}
# Generate observed SFS from the genomaps object
obs_sfs <- sfs(gm1001g)

# Plot the observed SFS
plot(obs_sfs)

# Expected (neutral) SFS from the genomaps object, which follows E(i) = theta/i
exp_sfs <- expsfs(gm1001g)
plot(exp_sfs)

# Tips: you can overlay the expected SFS on the observed SFS
plot(obs_sfs, expected = exp_sfs, log = 'x')

# Tips: you can also generate a folded SFS
obs_sfs_folded <- sfs(gm1001g, folded = TRUE)

# Tips: you can overlay the expected SFS by passing the genomaps object directly
# No need to run the expsfs() function first
plot(obs_sfs_folded, expected = gm1001g, log = 'x')
```

### 3b. Predict the MAR/GDAR using `MARtheory()`

`MARtheory()` predicts the MAR/GDAR analytically from the site frequency spectrum alone,
assuming individuals mate at random and are distributed at random in space. Under those two
assumptions, sampling a larger area is equivalent to sampling more individuals, and the
spectrum by itself determines how fast mutations accumulate. The function returns a
`martheory` data frame indexed by the number of sampled individuals `N`, so its output
lines up with `MARsampling()` in Step 4.

The *Arabidopsis* accessions satisfy neither assumption: they are selfing inbred lines, and
they are spatially structured across the native range. The prediction below consequently
overshoots the empirical curve, and that gap is informative rather than a failure of the
theory. `vignette('mar-theory')` takes the two assumptions apart one at a time and shows
how much each contributes to the fitted $z$.

```{r mar-theory}
theorydf <- MARtheory(gm1001g)
head(theorydf)

# plot the theorydf
plot(theorydf)

# you don't have to run MARcalc() separately: supply `fit = TRUE` and the plot method for
# the `martheory` class will fit and overlay the line for you
plot(theorydf, fit = TRUE)
```


## Step 4: Estimate the mutations-area relationship

The MAR is estimated in two steps.

### 4a. Generate the diversity-by-area sampling table using `MARsampling()`

`MARsampling()` builds the MAR/GDAR empirically. Working from the reference raster in the
`marmaps` object, it draws replicated square subsamples of the range, stepping from a
single raster cell up to the shorter side of the map, and computes the genetic diversity
inside each box. Boxes are placed under one of five schemes: `random`; the directional
`southnorth` and `northsouth`, which simulate climate-driven range shifts; and the focal
`outwards` and `inwards`, which simulate habitat destruction spreading from or towards the
most densely sampled cell. By default `nrep = 10` boxes are drawn at each size, box size
steps by `xfrac = 0.01` of the shorter map side, and `quorum = TRUE` discards boxes that
contain no occupied cell.

The returned `marsamp` data frame carries one row per box and four genetic diversity
metrics: the number of mutations `M`, endemic mutations `E`, Watterson's $\theta$ `thetaw`, and
nucleotide diversity `thetapi`. Area follows two strategies. `Asq` is the area of the whole
sampling square regardless of occupancy, and `A` sums only the occupied cells. `Asq`
accounts for variation in sample density across space, while `A` typically gives the
stronger correlation. The number of individuals in the box, `N`, is recorded as well, which
is what makes the table directly comparable with `MARtheory()`.

```{r mar-sampling}
mardf <- MARsampling(gm1001g, myseed = 91)
head(mardf)

# mardf is a marsamp object
class(mardf)

# plot the diversity-by-area sampling results
plot(mardf)

# Tips: alternative genetic diversity metrics or area metrics can be passed
plot(mardf, Mtype = 'M', Atype = 'Asq')
plot(mardf, Mtype = 'thetapi', Atype = 'A')
```

### 4b. Fit the mutations-area relationship model using `MARcalc()` {#4b}

`MARcalc()` fits the power-law model $M = cA^z$ using `sars::sar_power()` and returns the
fitted `sars` object. The scaling factor $z$ is the quantity of interest: it describes the
rate of genetic diversity loss and reflects the mutation abundance pattern, spatial structure, and breeding structure (see `vignette('mar-theory')` for details). `MARcalc()` accepts any of the `martheory`, `marsamp` or
`marextinct` objects, together with any combination of diversity metric (`Mtype`) and area
metric (`Atype`).

```{r mar-fit}
# print available Mtype and Atype
mar:::.Mtype
mar:::.Atype

marfit <- MARcalc(mardf)
# marfit is a sars object (from R package sars)
class(marfit)
# To extract full mar fit table, use summary
summary(marfit)

# Plot the fit
plot(mardf, fit = marfit$par)

# Alternatively, the plot.marsamp method can automatically fit the power-law model
plot(mardf, fit = TRUE)

# The same mutations fitted against the total square area instead
marfit_Asq <- MARcalc(mardf, Mtype = "M", Atype = "Asq")
summary(marfit_Asq)
plot(mardf, fit = marfit_Asq$par, Mtype = "M", Atype = "Asq")
```

`Asq` counts the empty space inside each box, so it captures uneven sampling density at the
cost of a weaker fit ($R^2$ `r round(summary(marfit)$R2, 2)` against `A`,
`r round(summary(marfit_Asq)$R2, 2)` against `Asq`). 
We recommend exploring both options for your data, especially if spatial sampling is uneven. 

The same machinery also fits the other diversity metrics, giving the broader GDAR.
Nucleotide diversity (`thetapi`) measures the average number of pairwise differences per
site between sampled sequences. Being dominated by common variants and largely insensitive
to rare ones, it saturates with area far faster than the number of mutations does, so its
$z$ is much shallower and its relationship noisier.

```{r mar-fit-pi}
marfit_pi <- MARcalc(mardf, Mtype = "thetapi")

# marfit is a sars object (from R package sars)
class(marfit_pi)
# To extract full mar fit table, use summary
summary(marfit_pi)

# Plot the fitted power-law model
plot(mardf, fit = marfit_pi$par, Mtype = "thetapi")
```

**Alternative workflow**:
    For alternative sampling schemes and other MARsampling options. See [Alternative 3: More on `MARsampling()`](#a3).

## Step 5: Estimate extinction and genetic diversity loss

The MAR framework can be applied in reverse to project the genetic diversity lost when
habitat disappears. `MARextinction()` removes raster cells iteratively, modelling local
extinctions where habitat becomes unavailable, and tracks the diversity remaining in the
surviving samples. It offers the same five schemes and the same four diversity metrics as
`MARsampling()`. Because individuals are removed directly from occupied cells and no
sampling square is ever constructed, however, area is reported only as `A`.

```{r mar-extinction}
extdf <- MARextinction(gm1001g, myseed = 91)
tail(extdf)

# extdf is a marextinct object
class(extdf)
```

Rearranging $M = cA^z$ gives the genetic extinction prediction. If a proportion
$a$ of the area is lost and a proportion $m$ of genetic diversity is lost with it, then
$M(1-m) = c[A(1-a)]^z$, which reduces to $1 - m = (1 - a)^z$. We can therefore take the
scaling factor fitted in
[4b. Fit the mutations-area relationship model using `MARcalc()`](#4b) and overlay its
prediction on the simulated extinction curve.

```{r mar-extinct-plot}
# plot the extinction mar fit
plot(extdf, fit = marfit$par)
```

`extdf` shares its structure with `mardf`, so it too can be passed to `MARcalc()`: sampling
and extinction are simply two ways of sampling the same spatial data. The $z$ they return
is not the same one.

```{r mar-extinct-fit}
extfit <- MARcalc(extdf)
plot(extdf, fit = extfit$par)
```

`MARextinction()` removes cells at random across the whole map, so the accessions left
behind approximate a random draw of genotypes. `MARsampling()` draws contiguous boxes
instead, and because *Arabidopsis* accessions are spatially structured, individuals within
one box are more closely related than a random draw would be (isolation-by-distance). A box
therefore captures fewer distinct mutations than an equally sized random sample, while both
converge on the same total across the full range. The sampled curve climbs more steeply as
a result, giving the higher $z$ of `r round(marfit$par['z'], 2)` against
`r round(extfit$par['z'], 2)` here.

**Alternative workflow**:
    For alternative sampling schemes and other MARextinction options. See [Alternative 4: More on `MARextinction()`](#a4).


## Step 6: Inspect and export outputs

Every object in the workflow is a plain data frame or list underneath its class, so `str()`,
`print()` and `summary()` work throughout. To obtain the $z$ values, run `summary()` on
the fitted `sars` object: it reports $c$ and $z$ with their confidence intervals along with
the fit diagnostics you need to judge whether the power law describes your data well.

```{r inspect}
# Object structures
str(extdf)
str(gm1001g)
str(marfit)

# Power-law model summaries
print(marfit)
# This generates the confidence interval
summary(marfit)
```

To save results or export to CSV (does not run):

```{r export, eval=FALSE}
save(gm1001g, file = 'gm1001g.rda')
saveRDS(marfit, file = "marfit.rds")
write.csv(mardf, file = "mardf.csv")
```


# The alternative `mar` workflow

## Alternative 1: Reading VCF files {#a1}

`vcf_parser()` reads a VCF file (`.vcf` or `.vcf.gz`) and returns a `margeno`
object, the same type of object returned by `text_parser()`. Use it in place
of [1a: Import genotype data using `text_parser()`](#s1) whenever your
genotype data is already in VCF format.

```{r import vcf}
# Tips: vcf_parser reads both gzipped and uncompressed VCF files.

# Locate example VCF file
vcffile <- system.file("extdata", "gmexp.vcf.gz", package = "mar")
# Tips: for your own data, change the variable above to the corresponding filepath

# Read genotype
genodata <- vcf_parser(vcffile, ploidy = 2)

print(genodata)
```

Sample coordinates are still imported separately with `lonlat_parser()`, as in
[1b: Import geographic data using `lonlat_parser()`](#s1b).

```{r import vcf coordinates}
# Locate example coordinates
lonlatfile <- system.file("extdata", "gmexp_lonlat.csv", package = "mar")
# Tips: this example uses arbitrary planar coordinates rather than longitude/latitude,
# so no coordinate reference system conversion is needed
mapsdata <- lonlat_parser(lonlatfile, incrs = "", mapcrs = "")

print(mapsdata)
```

With `genodata` and `mapsdata` in hand, proceed to Step 2 to construct the
`genomaps` object as usual.

## Alternative 2: Customizing projection and map resolution {#a2}

`lonlat_parser()` accepts three arguments that control how sample coordinates
are turned into a raster: `mapcrs`, `incrs`, and `mapres`.
[1b: Import geographic data using `lonlat_parser()`](#s1b) uses the defaults
for `incrs` and `mapres`; this section covers overriding them.

**Coordinates already in the target projection**: if your coordinates are not
longitude/latitude, or are already expressed in the same units as `mapcrs`,
set `incrs` to match `mapcrs` so that `lonlat_parser()` skips reprojection.

```{r skip reprojection}
# Locate example coordinates (arbitrary planar units, not longitude/latitude)
lonlatfile <- system.file("extdata", "gmexp_lonlat.csv", package = "mar")
# Tips: setting incrs = mapcrs (here, both "") skips reprojection entirely
mapsdata <- lonlat_parser(lonlatfile, incrs = "", mapcrs = "")

print(mapsdata)
```

**Custom map resolution**: by default `mapres = NULL`, so `mar` calculates
the resolution automatically following equation 12 in (Hengl 2006). Provide
`mapres` directly to override this, for example when you want raster cells to
match a specific spatial scale.

```{r custom resolution}
# Locate sample coordinates
lonlatfile <- system.file("extdata", "1001g_lonlat.txt", package = "mar")
# Tips: mapres is expressed in the units of mapcrs (meters, for EPSG:8857)
mapsdata <- lonlat_parser(lonlatfile, mapcrs = "EPSG:8857", mapres = 500000)

print(mapsdata)
plot(mapsdata)
```

## Alternative 3: More on `MARsampling()` {#a3}

The default `random` scheme places boxes uniformly across the map. The other four weight
placement using a beta distribution. `southnorth` and `northsouth` put boxes towards the southern or
northern edge, standing in for a climate-driven range shift, while `outwards` and `inwards`
put them away from or towards the most densely sampled cell, standing in for habitat
destruction radiating from a focal point. Comparing schemes is a quick way to see how
sensitive your $z$ is to the direction in which the range is sampled.

```{r sampling-schemes}
# a directional scheme, sampling from the south
mardf_sn <- MARsampling(gm1001g, scheme = "southnorth", myseed = 91)
plot(mardf_sn, fit = TRUE)

# the scheme used is stored as an attribute
attr(mardf_sn, "scheme")
```

`nrep` and `xfrac` trade resolution against runtime: `nrep` sets how many boxes are drawn
at each size, and `xfrac` sets the step between consecutive box sizes as a fraction of the
shorter map side. The defaults (`nrep = 10`, `xfrac = 0.01`) give a dense curve at some
cost, and coarsening them as above is the fastest way to explore a new dataset.

Setting `animate = TRUE` replays the sampling boxes on the reference map one at a time,
which is the quickest way to confirm that a scheme is doing what you expect. It outputs to
the default plotting device, so we do not evaluate it here.

```{r sampling-animate, eval=FALSE}
MARsampling(gm1001g, scheme = "outwards", nrep = 2, xfrac = 0.1, animate = TRUE, myseed = 91)
```

## Alternative 4: More on `MARextinction()` {#a4}

`MARextinction()` takes the same `scheme`, `nrep` and `xfrac` arguments, but they describe
removal rather than sampling. The scheme sets the order in which cells go extinct, `nrep`
sets how many independent extinction replicates are run, and `xfrac` sets the fraction of
occupied cells removed at each step. Each replicate ends with a row of zeros, representing
the loss of the whole range.

```{r extinction-schemes}
# extinction radiating outwards from the most densely sampled cell
extdf_out <- MARextinction(gm1001g, scheme = "outwards", nrep = 2, xfrac = 0.1, myseed = 91)
plot(extdf_out, fit = TRUE)
```

`animate = TRUE` likewise replays the extinction process on the map.

```{r extinction-animate, eval=FALSE}
MARextinction(gm1001g, scheme = "outwards", nrep = 2, xfrac = 0.1, animate = TRUE, myseed = 91)
```

# Troubleshooting

## Preparing input data

The `mar` package performs extensive checks on the input data to ensure that the results are reliable and interpretable.
Please make sure to check the `inst/extdata` folder for example files.

Here are the list of checks that are performed, and some tips on how to prepare the data:

### Genotype Data

- The pipeline requires bi-allelic SNP or invariant sites, so that the model assumptions hold.
  - Example `bcftools` command: `bcftools view -m2 -M2 -v snps ${VCF}`
- Missing data is allowed and accounted for. `mar` applies `pixy`'s site-level denominator and effective sequence length adjustments, which keep `thetaw` and `thetapi` unbiased in the presence of missing genotypes. Excessive missingness will still bias the summary, so we recommend imputing beforehand.
  - You can use tools like [beagle](https://faculty.washington.edu/browning/beagle/beagle.html) to impute missing data.
  - You can also filter the genotype data to retain only sites without missing data, e.g., `bcftools view -i 'N_MISSING == 0' ${VCF}`.
- It works best with diploid genotype data, but can handle any ploidy as long as it is consistent across all samples. Use caution when interpreting results for non-diploid data.
- If the reference genome is divergent from the species/population of interest, set the major allele as the reference allele to avoid issues with ancestral state identification.

### Geographic data

- Every sample with genotype data must have pairing longitude and latitude data.
- File format should be tab-delimited or comma-delimited with header (ID, LON/LONGITUDE, LAT/LATITUDE)
- Only three columns are allowed (ID, LON/LONGITUDE, LAT/LATITUDE)
- Sample IDs must be unique and in the same order as the Sample IDs provided in the genotype matrix.

