---
title: "The sampling theory behind MAR"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{The sampling theory behind MAR}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown:
        wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    fig.align = "center",
    fig.width = 6,
    fig.height = 4.5,
    dpi = 150,
    out.width = "100%"
)
rm(list = ls())
```

# Why does the mutations-area relationship exist?

In macroecology, species-area relationships (SAR) can be described exhaustively from
species abundance distributions (SAD) and species spatial distributions (Tjørve et al.
2008), and extensive studies have explored the theoretical relationship between the two
over the past century (Preston 1948, 1962; May 1975; Harte, Kinzig, and Green 1999;
Hubbell 2001). Extending this framework into population genetics, we show that a similar
sampling relationship exists between the site frequency spectrum (SFS), a fundamental
population genetics summary statistic, and the mutations-area relationship (MAR).

In this vignette, we first establish the theoretical connection between the SFS and the
SAD as scaled probability densities over abundance classes, and note one small
distinction. Then we introduce the `MARtheory()` function. Assuming that individuals are
**randomly mating and randomly distributed** in the population, `MARtheory()` derives the
exact mutations-area relationship from the site frequency spectrum alone, together with
the population density $\rho$, which links area to the number of sampled individuals,
$N = \rho A$ (Preston 1962). Without knowing any individual's specific location,
and without running the `MARsampling()` procedure, the SFS deterministically gives us the MAR. We
then validate `MARtheory()`'s prediction against `MARsampling()` and `MARextinction()` on a simulated dataset (the `gmexp` data).

Next, we show what happens when the two key assumptions of **random mating and random
spatial distribution** are violated in real life, using the *Arabidopsis thaliana* 1001
Genomes example (the `gm1001g` data): the theoretical prediction and the `MARsampling()`
output part ways. When we then treat the 1001 Genomes accessions as haploid individuals
and shuffle their sample locations, `MARtheory()` recovers the `MARsampling()` output in a
real dataset as well.

In the end, having shown that the MAR can be derived from the SFS, we discuss what the
$z$ value captures, which of the several $z$ values `mar` reports to carry forward into a
conservation prediction, and what that prediction does and does not cover.

TL,DR: directly skip to this [important interpretation section](#interpret).

## Some assumptions and the example data

We start by simulating 50 diploid individuals sampled from a randomly mating, randomly
distributed population growing exponentially, using the python package `msprime` (code
available in `data-raw/gmexp_simulate.py`). We chose a growing population, rather than a
constant-size one, to show that the theory holds regardless of the spectrum's shape.

We built a `genomaps` object from that
simulation (code available in `data-raw/gmexp_rda.R`) and provide it with this package.

You can explore the data here:

```{r load-gmexp}
library(mar)

data(gmexp)
print(gmexp)
```

To elaborate on the two key assumptions for deriving the MAR from an SFS:

**1. Individuals are randomly mating.** Under random mating an individual is just
`ploidy` independently drawn gene copies, so *sampling one diploid individual is the same
as sampling two random chromosomes*. This equivalence is what lets us use the SFS, which
does its bookkeeping at the chromosome level, while we actually collect individuals.

**2. Individuals are randomly distributed in space.** This is Preston's classical
assumption, and it is what allows area to stand in for sample size: if individuals are
scattered at random, a box twice as large holds twice as many of them, so $N \propto A$.
The `gmexp` coordinates were drawn from a uniform distribution, which we can see by
plotting the `marmaps` object stored within the `genomaps` object.

```{r plot-gmexp}
plot(gmexp$maps)
```

Both assumptions are essential for the derivation below. In the *Arabidopsis* section we
violate each one in turn and measure what it costs.

## What is a site frequency spectrum?

Before we get to the sampling relationship, we need to define what an SFS is.

Take a genetic dataset of $J$ individuals with ploidy $x$, there are $xJ$
chromosomes and $L$ biallelic SNP sites ($L \times xJ$ genotype matrix). At any one SNP site, count how many of
the $xJ$ chromosomes carry the mutation and call that count $k$. The **site frequency
spectrum** (SFS) is the tally of how many sites have $k = 0$ (invariants), how many have
$k = 1$ (singletons), how many have $k = 2$, and so on. We write the proportional site
frequency spectrum as $\sigma(k)$, so that $\sum_{k=0}^{xJ} \sigma(k) = 1$.

As an example, suppose we genotyped $L = 5$ biallelic sites across $xJ = 4$ chromosomes, with
allele counts $k = 1, 1, 3, 2, 1$ at the five sites. The raw SFS tallies these into a
vector indexed by $k$: three singletons, one doubleton, one tripleton, and no site fixed
for either allele, i.e. $(0, 3, 1, 1, 0)$ for $k = 0, \ldots, 4$. Dividing each entry by
$L$ gives the proportional spectrum, $\sigma(1) = 3/5$, $\sigma(2) = 1/5$, $\sigma(3) =
1/5$, and $\sigma(0) = \sigma(4) = 0$. `sfs()` below returns the raw counts; dividing by
$L$ recovers $\sigma(k)$.

For `gmexp` that is $L = `r length(gmexp$geno$allele_count)`$ sites spread over
$xJ = `r length(gmexp$maps$sample.id) * gmexp$geno$ploidy`$ chromosomes:

```{r sfs-gmexp}
obs <- sfs(gmexp)
plot(obs)
```

The mutations-area relationship arises from a core principle: the commonness of rarity.
Most mutations in a population are rare and only a few reach high frequencies, and the
shape of that spectrum is what gives rise to the
scaling behaviour we see in the MAR. This mirrors the well-known ecological pattern: just
as the species-area relationship emerges from the species abundance distribution, the
mutations-area relationship is a consequence of the site frequency spectrum, both rooted
in the statistical structure of rarity in biological systems.

There is, however, **a key distinction between the SFS and the SAD**. The SAD entries sum
to the total number of sampled individuals, whereas the SFS entries sum to the total
number of variant sites $L$, which is unrelated to how many individuals were sampled. Any
species-abundance model can be fitted to an SFS, since both are densities over abundance
classes, but they are not the same object. For a direct comparison to a SAD, a haplotype
abundance distribution under the infinite-alleles assumption would be more appropriate.

Excitingly, the mutations-area relationship emerges from *any* shape of the spectrum, as
long as individuals are randomly mating and randomly distributed. The population may be
under selection or may have passed through demographic change; we simply take the
spectrum as given. `gmexp` illustrates this well precisely because it is *not* a
constant-size neutral population: it grew exponentially, and exponential growth leaves an
excess of rare variants. We can check that against the neutral expectation the package
ships, `expsfs()`:

```{r expsfs-gmexp}
plot(obs, expected = gmexp)
# pass the `log = 'x'` option to inspect the singletons
plot(obs, expected = gmexp, log = 'x')
# pass the `log = 'y'` option to inspect the mutation abundance
# (graphics::plot() will warn about the zero entries)
plot(obs, expected = gmexp, log = 'y')
```

Singletons are more common than the constant-size neutral model predicts, and
high-frequency variants are depleted. This reflects the underlying demography of exponential growth, and it does not interfere with the sampling theory, as we show next.

## How to build the MAR from a site frequency spectrum?

### `MARtheory()` implements an analytical sampling result

Recall the two assumptions: individuals mate at random, so sampling $N$ individuals is
the same as sampling $xN$ random chromosomes; and individuals are distributed randomly in
space, so sampling a larger area means sampling more individuals.

Consider a single genomic site at which $k$ of the $xJ$ chromosomes carry the mutation,
and suppose we draw $N$ individuals, that is $xN$ chromosomes. We miss that mutation only
if every one of the $xN$ draws came from the $xJ - k$ chromosomes that do not carry it.
Counting the ways that can happen gives:

$$
P(\text{missed}) \;=\; \frac{\binom{xJ - k}{xN}}{\binom{xJ}{xN}},
$$

So we sample this mutation with probability $1 - P(\text{missed})$. A mutation carried by
half the population ($k = xJ/2$) is almost certain to be sampled, whereas a singleton
($k = 1$) is sampled only with probability $N/J$. Averaging this sample probability
across sites, weighted by how many sites carry each allele count $k$ ($\sigma(k)$), gives the expected
number of mutations in a sample of $N$ individuals. The full derivation is in [the math
section](#math).

We can therefore predict the whole mutations-area relationship with the equation below,
which is what `MARtheory()` implements:

$$
\boxed{\;E[M] \;=\; L \sum_{k=0}^{xJ} \sigma(k)
\left(1-\frac{\binom{xJ-k}{xN}}{\binom{xJ}{xN}}\right)\;}
$$

`MARtheory()` evaluates this purely from theory, without running any sampling scheme. It
reports its predictions against the number of sampled *individuals* $N$, in a column named
`N` so that the output lines up with `MARsampling()`, and we can fit the power law
$M = c N^z$ with `MARcalc()`.

```{r martheory-gmexp}
theorydf <- MARtheory(gmexp)
head(theorydf)

# plot the theorydf
plot(theorydf)

# you don't have to run MARcalc() separately: supply `fit = TRUE` and the plot method for
# the `martheory` class will fit and overlay the line for you
plot(theorydf, fit = TRUE)

# to expose the sar object the plot method builds internally, run MARcalc()
theorydf_fit <- MARcalc(theorydf, Mtype = 'M', Atype = 'N')
summary(theorydf_fit)

# overlay the fitted MARcalc results on the plot
plot(theorydf, fit = theorydf_fit$par)
```

### `MARsampling()` confirms the `MARtheory()` prediction

We can now compare the prediction against the sampling output, which builds the same
curve the hard way, by repeatedly sampling boxes on the map:

```{r marsamp-gmexp}
mardf <- MARsampling(gmexp, myseed = 1)
head(mardf)
# We need to set `Atype = 'N'` here since the MARtheory() is based on sample numbers not area size
plot(mardf, Mtype = "M", Atype = "N", theory = theorydf)
```

The blue prediction lands on top of the sampled points. When we also fit the scaling factor $z$ on
`MARsampling()` output using `MARcalc()`, the two `MARcalc()` outputs are very similar.

```{r marsamp-marcalc-gmexp}
# fit the MAR
mardf_fit <- MARcalc(mardf, Mtype = 'M', Atype = 'N')
summary(mardf_fit)

# you can compare the theory prediction with the empirical fit side by side
plot(mardf, Mtype = "M", Atype = "N", theory = theorydf, fit = mardf_fit$par)

# or use a log-log scale by passing graphics::plot the log option
plot(mardf, Mtype = "M", Atype = "N", theory = theorydf, fit = mardf_fit$par, log = 'xy')
```


### `MARextinction()` confirms the `MARtheory()` prediction

Because `MARextinction()` simulates a different sampling process than `MARsampling()`, it
is worth checking `MARtheory()`'s prediction against its output too. It builds the same
curve the hard way as well, by iteratively removing the available samples on the map:

```{r marext-gmexp}
extdf <- MARextinction(gmexp, myseed = 1)
head(extdf)

# the blue line lands on the extinction curve constructed
plot(extdf, Mtype = "M", Atype = "N", theory = theorydf)

# you can compare the theory prediction with the empirical fit side by side.
# notice we did not invoke MARcalc() as above, as the `fit = TRUE` option will automatically run MARcalc()
plot(extdf, Mtype = "M", Atype = "N", theory = theorydf, fit = TRUE)

# to expose the sar object the plot method builds internally, run MARcalc()
extdf_fit <-MARcalc(extdf, Mtype = 'M', Atype = 'N')
summary(extdf_fit)
```

Finally, let's put the three `MARcalc()` outputs side by side.
Notice that when our two assumptions of random mating and random spatial distribution are
met, the scaling factor $z$ is the same across all three functions' output, `MARtheory()`,
`MARsampling()` and `MARextinction()`, because they all describe the same sampling
relationship. Under these conditions, only information carried by the SFS contributes to $z$.

```{r three-mar-gmexp}
extract_summary <- function(fittedlist, names) {
    outdf <- data.frame()
    for (ii in seq_along(fittedlist)) {
        fitted <- fittedlist[[ii]]; name <- names[ii]
        fittedsum <- summary(fitted)$Parameters
        rownames(fittedsum) <- NULL
        output <- data.frame(sampling_func = name, vars = c('c', 'z'))
        output <- cbind(output, fittedsum)
        outdf <- rbind(outdf, output)
    }
    return(outdf)
}

extract_summary(list(theorydf_fit, mardf_fit, extdf_fit),
                c('MARtheory', 'MARsampling', 'MARextinction'))
```

### Predictions for Watterson's $\theta$ and nucleotide diversity

The number of mutations $M$ is one measure of genetic diversity, and the same sampling
mechanism predicts Watterson's $\theta$ ($\theta_w$) and nucleotide diversity
($\theta_\pi$) as well. These are the metrics that turn the mutations-area relationship
into the broader genetic-diversity-area relationship (GDAR), and they scale differently from one another.

**Watterson's $\theta$ increases with sample size here.** It counts sites that are
segregating among the $xN$ sampled chromosomes and divides by a sample-size correction,

$$
\hat{\theta}_w \;=\; \frac{S_{xN}}{L\,H_{xN-1}},
\qquad H_m \;=\; \sum_{i=1}^{m} \frac{1}{i},
$$

where $S_{xN}$ is the number of segregating sites in the subsample and $H_m$ is the $m$-th
harmonic number. The sampling argument above gives $E[S_{xN}]/L$ directly, with one extra
term: a site is *not* segregating if the subsample happens to draw all mutations (alternative alleles), which
happens with probability $\binom{k}{xN}/\binom{xJ}{xN}$. So

$$
E[\theta_w] \;=\; \frac{1}{H_{xN-1}} \sum_{k=0}^{xJ} \sigma(k)
\left(1-\frac{\binom{xJ-k}{xN}}{\binom{xJ}{xN}}-\frac{\binom{k}{xN}}{\binom{xJ}{xN}}\right).
$$

Under the neutral coalescent, $\theta_w$ is an unbiased estimator of the
population-scaled mutation rate, and it is constant regardless of $N$. But $H_{xN-1}$ is only the right
correction when the spectrum has the neutral $1/k$ shape. Our simulated `gmexp` grew
exponentially, its spectrum does not have that shape, and so the prediction climbs with
sample size, exactly as the sampled values do.

```{r thetaw-gmexp}
plot(mardf, Mtype = "thetaw", Atype = "N", theory = theorydf)
```

**Nucleotide diversity $\theta_\pi$ does not change with sample size.** It is the average
over all $\binom{xN}{2}$ pairs of sampled chromosomes of the per-site probability that the
pair differs,

$$
\hat{\theta}_\pi \;=\; \frac{1}{L} \sum_{i=1}^{L} \frac{2k_i(xN-k_i)}{xN(xN-1)},
$$

where $k_i$ is the allele count at site $i$ *within the subsample*. Any pair we draw is
just a uniformly random pair from the full dataset, so the pair-level expectation is the
full-dataset value no matter how many pairs we draw, and the prediction reduces to

$$
E[\theta_\pi] \;=\; \sum_{k=0}^{xJ} \sigma(k)\,\frac{2k(xJ-k)}{xJ(xJ-1)},
$$

in which $N$ does not appear at all. Subsampling therefore cannot shift it, for any
spectrum ([see the Math section for details](#math)).
```{r thetapi-gmexp}
plot(mardf, Mtype = "thetapi", Atype = "N", theory = theorydf)
```

### Going from sampling individuals to sampling areas

Everything so far predicts diversity as a function of sample size. Because the `gmexp`
individuals are randomly placed, sample size and area really are interchangeable, and the
regression of area on sample size closes the final link between the site frequency
spectrum and the mutations-area relationship:

```{r area-gmexp}
plot(A ~ N, data = mardf)
A_N_lm <- lm(A ~ N, data = mardf)
summary(A_N_lm)
# with N = rho * A, the fitted slope of A ~ N estimates 1 / rho
rho <- 1 / A_N_lm$coefficients['N']
print(rho)
```

We can therefore convert the predicted $N$ axis into an area axis and compare against the
sampled MAR.

```{r area-theory-gmexp}
# add the expected area
theorydf$A <- theorydf$N / rho

# overlay with the MARsampling output
plot(mardf)
lines(M ~ A, data = theorydf, col = 'blue')
```

We deliberately left this last step out of `MARtheory()`. In real-life practice, we are more interested in
the scaling value $z$ rather than the constant $c$ in the $M = cA^z$ relationship.
When the population density $\rho$ is a constant fitted from the `MARsampling()` output, the scaling factor $z$ for area is very similar to the scaling factor $z$ for sample size, as shown below:

```{r z-N-A}
mardf_fit_A <- MARcalc(mardf)
extract_summary(list(mardf_fit, mardf_fit_A), c('M ~ N', 'M ~ A'))
```

# How does this work in real life: our *Arabidopsis* dataset?

The *Arabidopsis thaliana* 1001 Genomes accessions break both assumptions at once, which
makes them a good case for demonstrating that the scaling factor $z$ is determined by
three things in a real, conservation-relevant setting:

1. the site frequency spectrum,
2. the breeding structure,
3. the spatial distribution.

```{r load-gm1001g}
data(gm1001g)
print(gm1001g)
```

Let us run the prediction and compare it against random subsamples of individuals, as we
did before.

```{r theory-broken-gm1001g}
theorydf_gm1001g <- MARtheory(gm1001g)
plot(theorydf_gm1001g, fit = TRUE)

mardf_gm1001g <- MARsampling(gm = gm1001g, myseed = 91)
plot(mardf_gm1001g, Atype = 'N', theory = theorydf_gm1001g)
```

The prediction overshoots badly, by tens of percent over most of the range, and the gap
closes only at the largest sample sizes.

## No longer randomly mating: the accessions behave as haploids

The genotypes explain why. *Arabidopsis thaliana* is a predominantly selfing species, and
these accessions are inbred lines that are *completely homozygous*:

```{r homozygous}
table(gm1001g$geno$genotype)
```

Not one heterozygote. We can see this more clearly in the SFS, where every odd allele count is empty.

```{r oddk}
sfs_gm1001g <- sfs(gm1001g, folded = FALSE)
plot(sfs_gm1001g, log = 'x')
```

This breaks the random mating assumption The two chromosomes inside an accession are not two
independent draws, they are identical copies. When we sample one inbred individual but assume random mating, we are assuming we randomly sampled two independent chromomes, so the theory predicts too many mutations.

The remedy is to describe the data as what it effectively is: a haploid sample of
`r length(gm1001g$maps$sample.id)` gene copies rather than a diploid sample of
`r 2 * length(gm1001g$maps$sample.id)`.

```{r gm1001gh}
# edit the genotype of gm1001g directly
gm1001gh_genotype <- gm1001g$geno$genotype
# manually create a "pseudo-haploid"
gm1001gh_genotype <- gm1001gh_genotype / 2
table(gm1001gh_genotype)
# construct a new margeno and genomaps object
gm1001gh_geno <- margeno(genotype = gm1001gh_genotype, ploidy = 1, sample.id = gm1001g$geno$sample.id)
gm1001gh <- genomaps(geno = gm1001gh_geno, maps = gm1001g$maps)

# run the theory prediction again
theorydf_gm1001gh <- MARtheory(gm1001gh)
plot(theorydf_gm1001gh, fit = TRUE)

# theorydf_gm1001gh is indexed by `N`, so we can reuse the original sampling output and
# overlay the "pseudo-haploid" prediction on it
plot(mardf_gm1001g, Atype = 'N', theory = theorydf_gm1001gh)
plot(mardf_gm1001g, Atype = 'N', theory = theorydf_gm1001gh, log = 'xy')
```

Describing the ploidy correctly removes most of the bias: the prediction now runs through
the sampled cloud instead of well above it. What is left is a predicted slope that is
shallower than the sampled points show, and that residual is spatial, which is the subject
of the next section.

Note that we "made up" a genotype here (and we will make up locations below) in order to
demonstrate the correctness of the sampling theory, because this *Arabidopsis* dataset is
an extreme case of inbreeding. For your own analysis we recommend caution in manipulating
the data unless you are very familiar with R and with your dataset. The theory overlay is
there to let you *see* how much the breeding structure moves your $z$ estimate, without
having to rewrite your data.

## No longer randomly distributed: sample size is not proportional to area

We have now repaired the bias between mutations and *sample size* introduced by the
breeding structure. The MAR is about *area*, though, and in real data the two are no
longer perfectly related:

```{r nagm1001g}
plot(A ~ N, data = mardf_gm1001g)
```

The scatter is now substantial, and different sampling schemes build different
$N \sim A$ relationships.
```{r nagm1001g-ext}
extdf_gm1001g <- MARextinction(gm1001g, myseed = 91)

# check the area - sample size relationship first
plot(A ~ N, data = extdf_gm1001g)

# overlay the theory prediction
plot(extdf_gm1001g, Atype = 'N', theory = theorydf_gm1001gh)
```

When we run `MARcalc()` on `MARextinction()` output in the *Arabidopsis* data, the fitted
$z$ value is no longer the same as the one from `MARsampling()`.

```{r ext-samp-gm1001g}
mardf_fit_gm1001g = MARcalc(mardf_gm1001g)
extdf_fit_gm1001g = MARcalc(extdf_gm1001g)
extract_summary(list(mardf_fit_gm1001g, extdf_fit_gm1001g),
                c('MARsampling', 'MARextinction'))
```

Note that the `MARtheory()` prediction tracks the *extinction* curve more closely than it
tracked the `MARsampling()` cloud. This is because `MARextinction()` randomly removes cells throughout the map iteratively, so the accessions that remain are close to a random sampling of the genotypes.
`MARsampling()` draws contiguous boxes instead, and because *Arabidopsis* accessions are spatially structured, the individuals inside one box are more related to each other than a random draw would be (isolation-by-distance), and they carry fewer distinct mutations than the theory expects.

To isolate the spatial effect, we can shuffle the sample locations so that the accessions
become "pseudo-" randomly distributed, and keep the genotype matrix as the original.

```{r gm1001gr}
set.seed(42)
nind <- length(gm1001g$maps$sample.id)
gm1001gr_loc <- data.frame(id = gm1001g$maps$sample.id,
                           lon = runif(nind, -10, 10),
                           lat = runif(nind, -10, 10))
# lon/lat drawn uniformly, then reprojected into an equal-area map
gm1001gr_maps <- marmaps(lonlatdf = gm1001gr_loc, mapcrs = "EPSG:8857")
gm1001gr <- genomaps(geno = gm1001g$geno, maps = gm1001gr_maps)

# check the new map we created
plot(gm1001gr$maps)
```

With the accessions scattered at random, area and sample size once again carry the same
information:

```{r gm1001gr-sample}
mardf_gm1001gr <- MARsampling(gm1001gr, myseed = 91)
plot(A ~ N, data = mardf_gm1001gr)
```

And now the haploid prediction recovers the sampled curve:

```{r zcompare}
plot(mardf_gm1001gr, Atype = 'N', theory = theorydf_gm1001gh)
```

Extinction simulations on the relocated dataset align with the same prediction.

```{r extgm1001gr}
extdf_gm1001gr <- MARextinction(gm1001gr, myseed = 91)
plot(extdf_gm1001gr, Atype = 'N', theory = theorydf_gm1001gh)
```

# The important take-away: what does the $z$ value capture? {#interpret}

Three separate things influence $z$ in a realistic dataset, and the sections above have
now identified them one by one.

1. **Mutation abundance**, summarized by the SFS. This is the part `MARtheory()`
   predicts, and in the randomly mating, randomly distributed `gmexp` example it accounts
   for the entire curve.
2. **Breeding structure.** The SFS counts chromosomes, but we sample individuals, and the
   two coincide only under random mating. Inbreeding pushes $z$ higher, as the selfing
   *Arabidopsis* accessions in `gm1001g` show.
3. **Spatial structure**, which enters twice: through population density, the $N \sim A$
   relationship, and through how related nearby individuals are. Random placement makes
   area a clean proxy for sample size and brings the area-based and size-based $z$ into
   agreement, whereas spatial distribution can move $z$ in either direction. Stronger 
   population structure and isolation-by-distance usually moves $z$ higher.

A $z$ estimated from real data contains information from all three aspects, which makes 
it an elegant and easy-to-interpret method to analyze genetic diversity at macroecological scale. 
One scaling factor on a percentage scale summarizes how genetic diversity is distributed on a landscape 
and how fast it would be lost if the area shrank. The `MARtheory()` function predicts a null $z$ value 
for a given data set, separating the contributions of the mutation abundance patterns from the breeding structure, and spatial structure.

## So which $z$ value do I use for conservation?

For one dataset and one genetic diversity metric, `mar` returns several $z$ values: from
the theoretical prediction (`MARtheory()`), from empirical sampling (`MARsampling()`), and
from extinction simulation (`MARextinction()`), each fitted against either occupied cell
area (`A`) or total square area (`Asq`). These estimates need not agree. In the `gm1001g`
fits above, `MARsampling()` gives $z = `r round(mardf_fit_gm1001g$par['z'], 2)`$ against
occupied cell area while `MARextinction()` gives
$z = `r round(extdf_fit_gm1001g$par['z'], 2)`$, and the haploid `MARtheory()` prediction
is shallower still.

They disagree because each samples the landscape in a different way, not because one of
them is wrong. We therefore recommend fitting both area strategies, inspecting the curves
from all three functions side by side. Finally, use the $z$ value from `MARsampling()` to predict 
genetic extinction with area loss. Alternatively, a range of $z$ values 
should be employed when a single estimate is hard to ascertain.

## What this $z$ does not cover

One caution belongs with every number produced this way. Everything in this vignette is a
sampling result: it describes the genetic diversity you would find in a smaller sample of
a population as it stands today. It is not a process model of mutation and drift, so it
speaks to the diversity lost the moment habitat disappears, not to the diversity that
erodes afterwards (Mualim et al. 2026).

# Math appendix: deriving the MAR from SFS {#math}

Throughout, let $J$ be the number of individuals in the full dataset, $x$ the ploidy, and
$xJ$ the total number of sequenced chromosomes. The genotype matrix covers $L$ sites. At
each site, $k$ denotes the number of chromosomes carrying the mutation, and $\sigma(k)$ is
the proportional site frequency spectrum, so that $\sum_{k=0}^{xJ}\sigma(k) = 1$. We
sample $N$ individuals, equivalently $xN$ chromosomes (this $N$ is what the package
reports in the `N` column), and write $H_m = \sum_{i=1}^{m} 1/i$ for the $m$-th harmonic
number.

The goal is the expected number of captured mutations $M$ as a function of $N$, which in
turn stands in for area.

## $xN = 1$ and $L = 1$

One chromosome is sequenced at a single site, where the mutation is carried by $k$ of the
$xJ$ chromosomes. $M$ can only be 0 or 1, so

$$
M \sim \operatorname{Bernoulli}\!\left(\frac{k}{xJ}\right).
$$

## $xN = 1$ and $L > 1$

With $L$ sites we have $L$ independent Bernoulli trials, but the success probability
differs from site to site because $k$ does. Treating $k$ as a random variable drawn from
the spectrum, $K \sim \sigma(k)$, the marginal success probability is

$$
\bar{p} \;=\; \sum_{k=0}^{xJ} \sigma(k)\,\frac{k}{xJ},
$$

and the total number of captured mutations is

$$
M \sim \operatorname{Binom}(L, \bar{p}).
$$

## $xN > 1$ and $L > 1$

When more than one chromosome is sampled, the per-site success probability is no longer
$k/xJ$; it is one minus a hypergeometric miss probability. Sampling $xN$ chromosomes
without replacement from $xJ$, the mutation is missed only when all $xN$ come from the
$xJ-k$ non-carriers:

$$
P(\text{captured}) \;=\; 1 - \frac{\binom{xJ-k}{xN}}{\binom{xJ}{xN}}.
$$

$M$ is still binomial, so its expectation is $L$ times the spectrum-averaged capture
probability:

$$
\boxed{\;E[M] \;=\; L \sum_{k=0}^{xJ} \sigma(k)
\left(1-\frac{\binom{xJ-k}{xN}}{\binom{xJ}{xN}}\right)\;}
$$

## The other two diversity measures

**Watterson's $\theta$** is defined on sites that are *segregating* in the subsample,
that is $1 \le K \le xN-1$. Both fixed outcomes must therefore be removed: the site can
come out monomorphic because no mutation was sampled, or because *only* mutations were
sampled. Dividing the resulting per-site fraction by $H_{xN-1}$ gives $E[\theta_w]$.
$E[\theta_w]$ is flat in $xN$ if and only if $\sigma(k) \propto 1/k$.
The neutral spectrum is a fixed point of hypergeometric projection. Using
$\frac{1}{k}\binom{k}{i} = \frac{1}{i}\binom{k-1}{i-1}$ and the Vandermonde convolution
$\sum_m \binom{m}{a}\binom{M-m}{b} = \binom{M+1}{a+b+1}$, a spectrum $\phi_k = \theta/k$
projects onto a subsample of $xN$ chromosomes as $\tilde\phi_i = \theta/i$ for
$i = 1,\dots,xN-1$. Summing gives $E[S_{xN}] = \theta H_{xN-1}$, so dividing by
$H_{xN-1}$ returns $\theta$ at every $xN$. For any other spectrum the cancellation is
incomplete. In general $\theta_w(xN)$ runs from $\theta_\pi$ at $xN = 2$ to the
full-dataset $\theta_w$ at $xN = xJ$, so the direction of its trend is the sign of $-D$,
where $D$ is Tajima's $D$.
An upward drift, as in `gmexp`, means the dataset carries an excess of rare variants.

$$
E[\theta_w] \;=\; \frac{1}{H_{xN-1}} \sum_{k=0}^{xJ} \sigma(k)
\left(1-\frac{\binom{xJ-k}{xN}}{\binom{xJ}{xN}}-\frac{\binom{k}{xN}}{\binom{xJ}{xN}}\right).
$$

**Nucleotide diversity** measures the average number of differences per site between DNA sequences.
$E[\theta_\pi]$ is invariant in $xN$, for any SFS. At a single
site, $\hat\theta_\pi = \Delta / \binom{xN}{2}$, where $\Delta$ counts discordant pairs
among the sampled chromosomes. By exchangeability of sampling without replacement, each of
the $\binom{xN}{2}$ sampled pairs is marginally a uniformly random pair drawn from the
$\binom{xJ}{2}$ pairs in the full dataset, so
$E[\Delta] = \binom{xN}{2}\,k(xJ-k)/\binom{xJ}{2}$. The $\binom{xN}{2}$ cancels on
division, leaving the full-dataset value for every $xN \ge 2$; averaging over sites with
weights $\sigma(k)$ gives the boxed expression. No model assumption enters. The argument
is pure combinatorics, which is why the $\theta_\pi$ curve stays flat regardless of
demography or selection.


$$
E[\theta_\pi] \;=\; \sum_{k=0}^{xJ} \sigma(k)\,\frac{2k(xJ-k)}{xJ(xJ-1)}
$$
