---
title: "MAR Analysis Workflow"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{MAR Analysis Workflow}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown:
        wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    fig.align = "center",
    fig.width = 6,
    fig.height = 4.5,
    dpi = 150,
    out.width = "100%"
)
rm(list = ls())
```

# The standard `mar` workflow

## Overview

This vignette demonstrates the stepwise workflow for conducting a Mutation-Area
Relationship (MAR) analysis using the `mar` package, illustrated with the included
 *Arabidopsis thaliana* demo data.

```{r setup}
library(mar)
```


## Step 0: Load example data

The `mar` package provides an example dataset in the `genomaps` data format ready for analysis.

```{r example}
data(gm1001g)
print(gm1001g)

# gm1001g is a list of two objects, to obtain marmaps and margeno objects
genodata <- gm1001g$geno
mapsdata <- gm1001g$maps

# Tips: all operations in `mar` package takes `genomaps` as the input.
# Therefore, you do not have to extract the marmaps and margeno objects most if not all the time.
```

If you do not need to load your own data, skip to [Step 3: Explore the Site Frequency Spectrum (SFS)](#s3).

Otherwise, continue to [Step 1: Import raw data](#s1)

## Step 1: Import raw data {#s1}

### 1a: Import genotype data using `text_parser()`

`text_parser()` reads the genotype matrix and returns a `margeno` object
containing the genotype data alongside sample and variant metadata.

TODO: Add file specific instructions in marApp.

```{r import genotype}
# Tips: Begin by locating and importing the bundled example data included with the
# `mar` package. `system.file()` resolves the path to files stored in `inst/extdata/`,
# ensuring the vignette works regardless of where the package is installed.

# Locate sample genotypes
# Tips: text_parser reads both gzipped and uncompressed text files.
genofile <- system.file("extdata", "1001g_genotypes.txt.gz", package = "mar")
# Optional: text_parser also reads sample names and chromosome positions if provided
samplefile <- system.file("extdata", "1001g_accessions.txt", package = "mar")
chrposfile <- system.file("extdata", "1001g_chrpos.txt", package = "mar")
# Tips: for your own data, change the variables above to the corresponding filepaths

# Read genotype
# Optional: for vcf parsing, see Alternative 1: Reading VCF files
genodata <- mar::text_parser(
    geno.fn = genofile,
    samp.fn = samplefile,
    pos.fn = chrposfile,
    ploidy = 2
)

print(genodata)
```

**Alternative workflow**:
    `vcf_parser()` handles reading of VCF files, see [Alternative 1: Reading VCF files](#a1).

### 1b: Import geographic data using `lonlat_parser()` {#s1b}

`lonlat_parser()` reads the coordinate file (TXT, CSV, or TSV) and returns a `marmaps` object.

TODO: this needs to be updated when `lonlat_parser()` returns `marmaps` object

TODO: fix the figure resolution problem (now it looks very blurry)

```{r import coordinates}
# Locate sample coordinates
lonlatfile <- system.file("extdata", "1001g_lonlat.txt", package = "mar")
# Read coordinates
# Tips: lonlat_parser will assume the coordinates are longitude and latitude following ESPG: 4326 (WGS84) by default
# and automatically determine the map resolution following equation 12 in (Hengl 2006)
mapsdata <- lonlat_parser(lonlatfile, mapcrs = "EPSG:8857")

print(mapsdata)
# Optional: to inspect the reference map created
plot(mapsdata)
```

**Alternative workflow**:
    To change map resolution and projection, see [Alternative 2: Customizing projection and map resolution](#a2).

## Step 2: Construct `genomaps` object using `genomaps()`

After successful importing of genotypes and geographic data, it is now time to construct the main object for `mar` workflow.

`genomaps()` combines the `margeno` and `marmaps` objects into a single analysis
object used by all downstream functions.

```{r genomaps}
gm <- genomaps(genodata, mapsdata)
save(gm, file = "tempgm.rda")
```


## Step 3: Explore the Site Frequency Spectrum using `sfs()` {#s3}

TODO: fill in the marApp description on SFS here.

The observed SFS summarises how allele counts are distributed across sites;
the expected SFS provides a neutral-theory baseline for comparison.

```{r sfs}
# Generate observed SFS from the genomaps object
obs_sfs <- sfs(gm1001g)

# Plot the observed SFS
plot(obs_sfs)

# Expected (neutral) SFS from the genomaps object, which follows E(i) = theta/i
exp_sfs <- expsfs(gm1001g)
plot(exp_sfs)

# Tips: you can also generate an unfolded SFS
obs_sfs_unfolded <- sfs(gm1001g, folded = FALSE)
plot(obs_sfs_unfolded)
exp_sfs_unfolded <- expsfs(gm1001g, folded = FALSE)
plot(exp_sfs_unfolded)
```

## Step 4: Estimate the mutations-area relationship

The MAR is estimated in two steps.

### 4a. Generate the diversity-by-area sampling table using `MARsampling()`

TODO: Description here on the output table. and the default options (e.g. nrep, xfrac). Use the marApp documentation or the manuscript when needed.

TODO: I cannot get the plot method to register for marsamp (crying in my heart >_<), could you check this?

```{r mar-animate}
MARsampling(gm1001g, myseed = 3, scheme = "outwards", animate = T)
```


```{r mar-sampling}
# Tips: uses of myseed not neccessary, used here for reproducibility with the manuscript
mardf <- MARsampling(gm1001g, myseed = 3, scheme = "outwards")
head(mardf)

# mardf is a marsamp object
class(mardf)
# plot the diversity-by-area sampling results (?mar::plot.marsamp)
mar:::plot.marsamp(mardf)
plot(mardf) # TODO: doesn't work but should work
```

### 4b. Fit the mutations-area relationship model using `MARcalc()` {#4b}

TODO: Description here

```{r mar-fit}
marfit <- MARcalc(mardf)
# marfit is a sars object (from R package sars)
class(marfit)
# To extract full mar fit table, use summary
summary(marfit)
# Extract the c and z values from marfit
c_M_A <- unname(marfit$par["c"])
z_M_A <- unname(marfit$par["z"])
# Plot the fit again
plot(mardf, fit = c(c = c_M_A, z = z_M_A))
```

You can also fit other mutation relationships. For example, nucleotide diversity thetapi

TODO: description here. What does theta_pi measures.

```{r}
# plot alternative diversity-by-area sampling results
mar:::plot.marsamp(mardf, Mtype = "thetapi", Atype = "Asq")
marfit_pi <- MARcalc(mardf, Mtype = "thetapi", Atype = "Asq")

# marfit is a sars object (from R package sars)
class(marfit_pi)
# To extract full mar fit table, use summary
summary(marfit_pi)
# Extract the c and z values from marfit_pi
c_pi_Asq <- unname(marfit_pi$par["c"])
z_pi_Asq <- unname(marfit_pi$par["z"])
# Plot the fit again
plot(mardf, fit = c(c = c_pi_Asq, z = z_pi_Asq), Mtype = "thetapi", Atype = "Asq")
```

**Alternative workflow**:
    For alternative sampling schemes and other MARsampling options. See [Alternative 3: More on `MARsampling()`](#a3).

## Step 5: Estimate Extinction / Diversity Loss

`MARextinction()` projects the expected loss of genetic diversity under
hypothetical area reduction scenarios.

```{r mar-extinction}
marext <- MARextinction(gm1001g, myseed = 91)
head(marext)

# marext is a marextinct object
class(marext)
# plot the diversity-by-area sampling results (?mar::plot.marsamp)
mar:::plot.marextinct(marext)
plot(marext) # TODO: doesn't work but should work
```

We can use the scaling factors inferred in [4b. Fit the mutations-area relationship model using `MARcalc()`](#4b) to predict the extinction.

```{r }
# plot the extinction mar fit
plot.marextinct(marext, z = z_M_A)
```

**Alternative workflow**:
    For alternative sampling schemes and other MARextinction options. See [Alternative 4: More on `MARextinction()`](#a4).


## Step 7: Inspect and Export Outputs

Use `str()`, `print()`, and `summary()` to inspect any result object.

```{r inspect}
# Object structures
str(mar_sampling)
str(mut_div)
str(mar_fit)
str(mar_ext)

# Model summaries
print(mar_fit)
summary(mar_fit)
```

To save results or export to CSV:

```{r export}
saveRDS(mar_fit, file = "mar_fit.rds")
write.csv(as.data.frame(mar_sampling), file = "mar_sampling.csv", row.names = FALSE)
```


# The alternative `mar` workflow

## Alternative 1: Reading VCF files {#a1}

`vcf_parser()` reads a VCF file (`.vcf` or `.vcf.gz`) and returns a `margeno`
object, the same type of object returned by `text_parser()`. Use it in place
of [1a: Import genotype data using `text_parser()`](#s1) whenever your
genotype data is already in VCF format.

```{r import vcf}
# Tips: vcf_parser reads both gzipped and uncompressed VCF files.

# Locate example VCF file
vcffile <- system.file("extdata", "gmexp.vcf.gz", package = "mar")
# Tips: for your own data, change the variable above to the corresponding filepath

# Read genotype
genodata <- mar::vcf_parser(vcffile, ploidy = 2)

print(genodata)
```

Sample coordinates are still imported separately with `lonlat_parser()`, as in
[1b: Import geographic data using `lonlat_parser()`](#s1b).

```{r import vcf coordinates}
# Locate example coordinates
lonlatfile <- system.file("extdata", "gmexp_lonlat.csv", package = "mar")
# Tips: this example uses arbitrary planar coordinates rather than longitude/latitude,
# so no coordinate reference system conversion is needed
mapsdata <- lonlat_parser(lonlatfile, incrs = "", mapcrs = "")

print(mapsdata)
```

With `genodata` and `mapsdata` in hand, proceed to Step 2 to construct the
`genomaps` object as usual.

## Alternative 2: Customizing projection and map resolution {#a2}

`lonlat_parser()` accepts three arguments that control how sample coordinates
are turned into a raster: `mapcrs`, `incrs`, and `mapres`.
[1b: Import geographic data using `lonlat_parser()`](#s1b) uses the defaults
for `incrs` and `mapres`; this section covers overriding them.

**Coordinates already in the target projection**: if your coordinates are not
longitude/latitude, or are already expressed in the same units as `mapcrs`,
set `incrs` to match `mapcrs` so that `lonlat_parser()` skips reprojection.

```{r skip reprojection}
# Locate example coordinates (arbitrary planar units, not longitude/latitude)
lonlatfile <- system.file("extdata", "gmexp_lonlat.csv", package = "mar")
# Tips: setting incrs = mapcrs (here, both "") skips reprojection entirely
mapsdata <- lonlat_parser(lonlatfile, incrs = "", mapcrs = "")

print(mapsdata)
```

**Custom map resolution**: by default `mapres = NULL`, so `mar` calculates
the resolution automatically following equation 12 in (Hengl 2006). Provide
`mapres` directly to override this, for example when you want raster cells to
match a specific spatial scale.

```{r custom resolution}
# Locate sample coordinates
lonlatfile <- system.file("extdata", "1001g_lonlat.txt", package = "mar")
# Tips: mapres is expressed in the units of mapcrs (meters, for EPSG:8857)
mapsdata <- lonlat_parser(lonlatfile, mapcrs = "EPSG:8857", mapres = 100000)

print(mapsdata)
plot(mapsdata)
```

## Alternative 3: More on `MARsampling()` {#a3}

TODO: mention the animate option

## Alternative 4: More on `MARextinction()` {#a4}

TODO: mention the animate option without evaluating

# Troubleshooting


