---
title: "The sampling theory behind MAR"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{The sampling theory behind MAR}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown:
        wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    fig.align = "center",
    fig.width = 6,
    fig.height = 4.5,
    dpi = 150,
    out.width = "100%"
)
rm(list = ls())
```

# Why does the mutations-area relationship exist?

In macroecology, species-area relationships (SAR) can be described exhaustively from
species abundance distributions (SAD) and species spatial distributions (Tjørve et al.
2008), and extensive studies have explored the theoretical relationship between the two
over the past century (Preston 1948, 1962; May 1975; Harte, Kinzig, and Green 1999;
Hubbell 2001). Extending this framework into population genetics, we show that a similar
sampling relationship exists between the site frequency spectrum (SFS), a fundamental
population genetics summary statistic, and the mutations-area relationship (MAR).

In this vignette, we first establish the theoretical connection between the SFS and the
SAD as scaled probability densities over abundance classes, and note one small
distinction. Then we introduce the `MARtheory()` function. Assuming that individuals are
**randomly mating and randomly distributed** in the population, `MARtheory()` derives the
exact mutations-area relationship from the site frequency spectrum alone, together with
the population density $\rho$, which links the number of sampled individuals to area, $N = \rho A$ (Preston 1962). Without knowing any individual's specific location,
and without running the `MARsampling()` procedure, the SFS deterministically gives us the MAR. We
then validate `MARtheory()`'s prediction against `MARsampling()` and `MARextinction()` on a simulated dataset (the `gmexp` data).

Next, we show what happens when the two key assumptions of **random mating and random
spatial distribution** are violated in real life, using the *Arabidopsis thaliana* 1001
Genomes example (the `gm1001g` data): the theoretical prediction and the `MARsampling()`
output part ways. When we then treat the 1001 Genomes accessions as haploid individuals
and shuffle their sample locations, `MARtheory()` recovers the `MARsampling()` output in a
real dataset as well.

In the end, having shown that the MAR can be derived from the SFS, we discuss what that
implies for interpreting the $z$ value in conservation genetics.

## Some assumptions and the example data

We start by simulating 50 diploid individuals sampled from a randomly mating, randomly
distributed population growing exponentially, using the python package `msprime` (code
available in `data-raw/gmexp_simulate.py`). We built a `genomaps` object from that
simulation (code available in `data-raw/gmexp_rda.R`) and provide it with this package.

You can explore the data here:

```{r explore-gmexp}
library(mar)

data(gmexp)
print(gmexp)
```

To elaborate on the two key assumptions for deriving the MAR from an SFS:

**1. Individuals are randomly mating.** Under random mating an individual is just
`ploidy` independently drawn gene copies, so *sampling one diploid individual is the same
as sampling two random chromosomes*. This equivalence is what lets us use the SFS, which
does its bookkeeping at the chromosome level, while we actually collect individuals.

**2. Individuals are randomly distributed in space.** This is Preston's classical
assumption, and it is what allows area to stand in for sample size: if individuals are
scattered at random, a box twice as large holds twice as many of them, so $N \propto A$.
The `gmexp` coordinates were drawn from a uniform distribution, which we can see by
plotting the `marmaps` object stored within the `genomaps` object.

```{r randomloc}
plot(gmexp$maps)
```

Both assumptions are essential for the derivation below. In the *Arabidopsis* section we
violate each one in turn and measure what it costs.

## What is a site frequency spectrum?

Before we get to the sampling relationship, we need to define what an SFS is and what
notation we will use for it.

Take a genetic dataset of $N$ individuals with ploidy $x$, so that there are $xN$
chromosomes in total, genotyped, and contains $L$ biallelic SNP sites. At any one SNP site, count how many of
the $xN$ chromosomes carry the mutation and call that count $k$. The **site frequency
spectrum** (SFS) is the tally of how many sites have $k = 1$ (singletons), how many have
$k = 2$, and so on. We write the proportional site frequency spectrum as $\sigma(k)$, so
that $\sum_{k=0}^{xN} \sigma(k) = 1$.
TODO: what is \sigma(k). Give an example that generate SFS, and get sigma(k) by dividing with L.

For `gmexp` that is $L = `r length(gmexp$geno$allele_count)`$ sites spread over
$xN = `r length(gmexp$maps$sample.id) * gmexp$geno$ploidy`$ chromosomes:

```{r sfsplot}
obs <- sfs(gmexp)
plot(obs)
```

The mutations-area relationship arises from a core principle: the commonness of rarity.
Most mutations in a population are rare and only a few reach high frequencies.
The site frequency spectrum (SFS) captures this distribution of allele frequencies and is a fundamental summary statistic in population genetics. It represents the number of mutations observed at different frequencies within a sample, and its shape is what gives rise to the
scaling behaviour we see in the MAR. This mirrors the well-known ecological pattern: just
as the species-area relationship emerges from the species abundance distribution, the
mutations-area relationship is a consequence of the site frequency spectrum, both rooted
in the statistical structure of rarity in biological systems.

There is, however, **a key distinction between the SFS and the SAD**. The SAD entries sum
to the total number of sampled individuals, whereas the SFS entries sum to the total
number of sites $L$, which is unrelated to how many individuals were sampled. Any
species-abundance model can be fitted to an SFS, since both are densities over abundance
classes, but they are not the same object. For a direct comparison to a SAD, a haplotype
abundance distribution under the infinite-alleles assumption would be more appropriate.

Excitingly, the mutations-area relationship emerges from *any* shape of the spectrum, as
long as individuals are randomly mating and randomly distributed. The population may be
under selection or may have passed through demographic change; we simply take the
spectrum as given. `gmexp` illustrates this well precisely because it is *not* a
constant-size neutral population: it grew exponentially, and exponential growth leaves an
excess of rare variants. We can check that against the neutral expectation the package
ships, `expsfs()`:

```{r neutral}
plot(obs, expected = gmexp)
# pass the `log = 'x'` option to inspect the singletons
plot(obs, expected = gmexp, log = 'x')
# pass the `log = 'y'` option to inspect the mutation abundance
# (graphics::plot() will warn about the zero entries)
plot(obs, expected = gmexp, log = 'y')
```

Singletons are more common than the constant-size neutral model predicts, and
high-frequency variants are depleted. That is exponential growth showing up in the
spectrum, and it does not interfere with the sampling theory, as we show next.

## How to build the MAR from a site frequency spectrum?

### `MARtheory()` implements an analytical sampling result

Recall the two assumptions: individuals mate at random, so sampling $n$ chromosomes is
the same as sampling $j = n/x$ individuals; and individuals are distributed randomly in
space, so sampling a larger area means sampling more individuals.

Consider a single genomic site at which $k$ of the $xN$ chromosomes carry the mutation,
and suppose we draw $n$ chromosomes. We miss that mutation only if every one of the $n$
draws came from the $xN - k$ chromosomes that do not carry it. Counting the ways that can
happen gives

$$
P(\text{missed}) \;=\; \frac{\binom{xN - k}{n}}{\binom{xN}{n}},
$$

so we capture the site with probability $1 - P(\text{missed})$. A mutation carried by
half the population ($k = xN/2$) is almost certain to be sampled, whereas a singleton
($k = 1$) is sampled only with probability $n/xN$. Averaging this sample probability
across sites, weighted by how many sites carry each allele count $k$, gives the expected
number of mutations in a sample of $n$ chromosomes. The full derivation is in [the math
section](#math).

We can therefore predict the whole mutations-area relationship with the equation below,
which is what `MARtheory()` implements:

$$
\boxed{\;E[M] \;=\; L \sum_{k=0}^{xN} \sigma(k)
\left(1-\frac{\binom{xN-k}{n}}{\binom{xN}{n}}\right)\;}
$$

`MARtheory()` evaluates this purely from theory, without running any sampling scheme. It
reports its predictions against the number of *individuals* $j = n/x$, in a column named
`N` so that the output lines up with `MARsampling()`, and we can fit the power law
$M = c N^z$ with `MARcalc()`.

```{r marexp}
theorydf <- MARtheory(gmexp)
head(theorydf)

# plot the theorydf
plot(theorydf)

# you don't have to run MARcalc() separately: supply `fit = TRUE` and the plot method for
# the `martheory` class will fit and overlay the line for you
plot(theorydf, fit = TRUE)

# to expose the sar object the plot method builds internally, run MARcalc()
theorydf_fit <- MARcalc(theorydf, Mtype = 'M', Atype = 'N')
summary(theorydf_fit)

# overlay the fitted MARcalc results on the plot
plot(theorydf, fit = theorydf_fit$par)
```

### `MARsampling()` confirms the `MARtheory()` prediction

We can now compare the prediction against the sampling output, which builds the same
curve the hard way, by repeatedly sampling boxes on the map:

```{r marexp-compare}
mardf <- MARsampling(gmexp, myseed = 1)
head(mardf)
plot(mardf, Mtype = "M", Atype = "N", theory = theorydf)
```

The blue prediction lands on top of the sampled points, and the scaling factor $z$ from
`MARsampling()` is very close to the one from `MARtheory()`.

```{r zexp}
# fit the MAR
mardf_fit <- MARcalc(mardf, Mtype = 'M', Atype = 'N')

summary(mardf_fit)

# extract the fitted c and z values
print(mardf_fit$par)
print(theorydf_fit$par)

# you can compare the theory prediction with the empirical fit side by side
plot(mardf, Mtype = "M", Atype = "N", theory = theorydf, fit = mardf_fit$par)

# or use a log-log scale by passing graphics::plot the log option
plot(mardf, Mtype = "M", Atype = "N", theory = theorydf, fit = mardf_fit$par, log = 'xy')
```

### `MARextinction()` confirms the `MARtheory()` prediction

Because `MARextinction()` simulates a different sampling process than `MARsampling()`, it
is worth checking `MARtheory()`'s prediction against its output too. It builds the same
curve the hard way as well, by iteratively removing the available samples on the map:

```{r marext-compare}
extdf <- MARextinction(gmexp, myseed = 1)
head(extdf)

# the blue line lands on the extinction curve constructed
plot(extdf, Mtype = "M", Atype = "N", theory = theorydf)

# you can compare the theory prediction with the empirical fit side by side.
# notice we did not invoke MARcalc() as above, but simply asked the plot method to fit
plot(extdf, Mtype = "M", Atype = "N", theory = theorydf, fit = TRUE)
```

Notice that when our two assumptions of random mating and random spatial distribution are
met, the scaling factor $z$ is the same across all three functions, `MARtheory()`,
`MARsampling()` and `MARextinction()`, because they all describe the same sampling
relationship. Under these conditions, only information carried by the SFS contributes to $z$.

### Predictions for Watterson's $\theta$ and nucleotide diversity

The same sampling mechanism predicts Watterson's $\theta$ ($\theta_w$) and nucleotide
diversity ($\theta_\pi$) as well.

**Watterson's $\theta$ increases with sample size here.** It counts sites that are
segregating among the $n$ sampled chromosomes and divides by a sample-size correction,

$$
\hat{\theta}_w \;=\; \frac{S_n}{L\,H_{n-1}},
\qquad H_m \;=\; \sum_{i=1}^{m} \frac{1}{i},
$$

where $S_n$ is the number of segregating sites in the subsample and $H_m$ is the $m$-th
harmonic number. The sampling argument above gives $E[S_n]/L$ directly, with one extra
term: a site is *not* segregating if the subsample happens to draw all mutations (alternative alleles), which
happens with probability $\binom{k}{n}/\binom{xN}{n}$. So

$$
E[\theta_w] \;=\; \frac{1}{H_{n-1}} \sum_{k=0}^{xN} \sigma(k)
\left(1-\frac{\binom{xN-k}{n}}{\binom{xN}{n}}-\frac{\binom{k}{n}}{\binom{xN}{n}}\right).
$$

Under the neutral coalescent, $\theta_w$ is an unbiased estimator of the
population-scaled mutation rate, and it is constant regardless of $n$. But $H_{n-1}$ is only the right
correction when the spectrum has the neutral $1/k$ shape. Our simulated `gmexp` grew
exponentially, its spectrum does not have that shape, and so the prediction climbs with
sample size, exactly as the sampled values do.

```{r theta_w}
plot(mardf, Mtype = "thetaw", Atype = "N", theory = theorydf)
```

**Nucleotide diversity $\theta_\pi$ does not change with sample size.** It is the average
over all $\binom{n}{2}$ pairs of sampled chromosomes of the per-site probability that the
pair differs,

$$
\hat{\theta}_\pi \;=\; \frac{1}{L} \sum_{i=1}^{L} \frac{2k_i(n-k_i)}{n(n-1)},
$$

where $k_i$ is the allele count at site $i$ *within the subsample*. Any pair we draw is
just a uniformly random pair from the full dataset, so the pair-level expectation is the
full-dataset value no matter how many pairs we draw, and the prediction reduces to

$$
E[\theta_\pi] \;=\; \sum_{k=0}^{xN} \sigma(k)\,\frac{2k(xN-k)}{xN(xN-1)},
$$

in which $n$ does not appear at all. Subsampling therefore cannot shift it, for any
spectrum ([Proposition 1](#math)).

```{r thetapi}
plot(mardf, Mtype = "thetapi", Atype = "N", theory = theorydf)
```

### Going from sampling individuals to sampling areas

Everything so far predicts diversity as a function of sample size. Because the `gmexp`
individuals are randomly placed, sample size and area really are interchangeable, and the
regression of area on sample size is essentially noiseless:

```{r nagmexp}
plot(A ~ N, data = mardf)
N_A_lm <- lm(A ~ N, data = mardf)
summary(N_A_lm)
# with N = rho * A, the fitted slope of A ~ N estimates 1 / rho
rho <- 1 / N_A_lm$coefficients['N']
print(rho)
```

We can therefore convert the predicted $N$ axis into an area axis and compare against the
sampled MAR. We deliberately left this last step out of `MARtheory()`, because it needs
the density $\rho$, which has to be fitted from the `MARsampling()` output. Note that the
prediction is still analytical: we used only the population density from
`MARsampling()`, and none of its sampled mutation or diversity values.

```{r add-area}
# add the expected area
theorydf$A <- theorydf$N / rho

# overlay with the MARsampling output
plot(mardf)
lines(M ~ A, data = theorydf, col = 'blue')
```

# How does this work in real life: our *Arabidopsis* dataset?

The *Arabidopsis thaliana* 1001 Genomes accessions break both assumptions at once, which
makes them a good case for demonstrating that the scaling factor $z$ is determined by
three things in a real, conservation-relevant setting:

1. the site frequency spectrum,
2. the breeding structure,
3. the spatial distribution.

```{r loadgm1001g}
data(gm1001g)
print(gm1001g)
```

Let us run the prediction and compare it against random subsamples of individuals, as we
did before.

```{r gm1001gbroken}
theorydf_gm1001g <- MARtheory(gm1001g)
plot(theorydf_gm1001g, fit = TRUE)

mardf_gm1001g <- MARsampling(gm = gm1001g, myseed = 1)
plot(mardf_gm1001g, Atype = 'N', theory = theorydf_gm1001g)
```

The prediction overshoots badly, by tens of percent over most of the range, and the gap
closes only at the largest sample sizes.

## No longer randomly mating: the accessions behave as haploids

The genotypes explain why. These accessions are inbred lines, and they are *completely
homozygous*:

```{r homozygous}
table(gm1001g$geno$genotype)
```

Not one heterozygote. The same fact shows up in the spectrum as a striking regularity:
every odd allele count is empty, because a mutation can only ever be carried in pairs.

```{r oddk}
sfs_gm1001g <- sfs(gm1001g, folded = FALSE)
plot(sfs_gm1001g, log = 'x')
```

This breaks assumption 1 outright. The two chromosomes inside an accession are not two
independent draws, they are identical copies. When we sample one individual and hand the
theory two chromosomes, it believes it had two independent chances to catch each mutation
when it really had one, so it predicts too many.

The remedy is to describe the data as what it effectively is: a haploid sample of
`r length(gm1001g$maps$sample.id)` gene copies rather than a diploid sample of
`r 2 * length(gm1001g$maps$sample.id)`.

```{r haploid}
# edit the genotype of gm1001g directly
gm1001gh_genotype <- gm1001g$geno$genotype
# manually create a "pseudo-haploid"
gm1001gh_genotype <- gm1001gh_genotype / 2
table(gm1001gh_genotype)
# construct a new margeno and genomaps object
gm1001gh_geno <- margeno(genotype = gm1001gh_genotype, ploidy = 1, sample.id = gm1001g$geno$sample.id)
gm1001gh <- genomaps(geno = gm1001gh_geno, maps = gm1001g$maps)

# run the theory prediction again
theorydf_gm1001gh <- MARtheory(gm1001gh)
plot(theorydf_gm1001gh, fit = TRUE)

# theorydf_gm1001gh is indexed by `N`, so we can reuse the original sampling output and
# overlay the "pseudo-haploid" prediction on it
plot(mardf_gm1001g, Atype = 'N', theory = theorydf_gm1001gh)
plot(mardf_gm1001g, Atype = 'N', theory = theorydf_gm1001gh, log = 'xy')
```

Describing the ploidy correctly removes most of the bias: the prediction now runs through
the sampled cloud instead of well above it. What is left is a predicted slope that is
shallower than the sampled points show, and that residual is spatial, which is the subject
of the next section.

Note that we "made up" a genotype here (and we will make up locations below) in order to
demonstrate the correctness of the sampling theory, because this *Arabidopsis* dataset is
an extreme case of inbreeding. For your own analysis we recommend caution in manipulating
the data unless you are very familiar with R and with your dataset. The theory overlay is
there to let you *see* how much the breeding structure moves your $z$ estimate, without
having to rewrite your data.

## No longer randomly distributed: sample size is not proportional to area

We have now repaired the bias between mutations and *sample size* introduced by the
breeding structure. The MAR is about *area*, though, and in real data the two are no
longer perfectly related:

```{r nagm1001g}
plot(A ~ N, data = mardf_gm1001g)
N_A_lm <- lm(A ~ N, data = mardf_gm1001g)
summary(N_A_lm)
```

The scatter is now substantial, and different sampling schemes trace out different
$N$--$A$ relationships and therefore different fitted $z$ values:

```{r nagm1001g-ext}
extdf_gm1001g <- MARextinction(gm1001g, myseed = 1)
plot(A ~ N, data = extdf_gm1001g)
N_A_lm_ext <- lm(A ~ N, data = extdf_gm1001g)
summary(N_A_lm_ext)

plot(extdf_gm1001g, Atype = 'N', theory = theorydf_gm1001gh)
```

Note that the haploid prediction tracks the *extinction* curve much more closely than it
tracked the `MARsampling()` cloud. The reason is not that $N$ and $A$ are better
correlated here, but that `MARextinction()` removes cells from across the whole map, so
the accessions that remain are close to a random subset of the panel. `MARsampling()`
draws contiguous boxes instead, and because *Arabidopsis* accessions are spatially
structured, the individuals inside one box are more related to each other than a random
draw would be, and they carry fewer distinct mutations than the theory expects.

To isolate the spatial effect, we can shuffle the sample locations so that the accessions
become "pseudo-" randomly distributed.

```{r relocate}
set.seed(42)
nind <- length(gm1001g$maps$sample.id)
gm1001gr_loc <- data.frame(id = gm1001g$maps$sample.id,
                           lon = runif(nind, -10, 10),
                           lat = runif(nind, -10, 10))
# lon/lat drawn uniformly, then reprojected into an equal-area map
gm1001gr_maps <- marmaps(lonlatdf = gm1001gr_loc, mapcrs = "EPSG:8857")
gm1001gr <- genomaps(geno = gm1001g$geno, maps = gm1001gr_maps)

# check the new map we created
plot(gm1001gr$maps)
```

With the accessions scattered at random, area and sample size once again carry the same
information:

```{r gm1001gr-sample}
mardf_gm1001gr <- MARsampling(gm1001gr, myseed = 1)
plot(A ~ N, data = mardf_gm1001gr)
N_A_lm_r <- lm(A ~ N, data = mardf_gm1001gr)
summary(N_A_lm_r)
```

And now the haploid prediction recovers the sampled curve: the exponent stops depending on
whether we regress against area or against sample size, and both agree with the theory.

```{r zcompare}
plot(mardf_gm1001gr, Atype = 'N', theory = theorydf_gm1001gh)
```

Extinction simulations on the relocated dataset align with the same prediction, as they
already did on the original map.

```{r extgm1001gr}
extdf_gm1001gr <- MARextinction(gm1001gr, myseed = 1)
plot(extdf_gm1001gr, Atype = 'N', theory = theorydf_gm1001gh)
```

Collecting the fitted exponents makes the pattern easy to read. `z_N` is fitted against
sample size, `z_A` against area, and `z_theory` is the single prediction from the
"pseudo-haploid" SFS:

```{r ztable}
zof <- function(df, Atype) MARcalc(df, Mtype = 'M', Atype = Atype)$par[['z']]
ztab <- data.frame(
    data = c("gm1001g", "gm1001g", "gm1001gr", "gm1001gr"),
    scheme = c("MARsampling", "MARextinction", "MARsampling", "MARextinction"),
    z_N = c(zof(mardf_gm1001g, 'N'), zof(extdf_gm1001g, 'N'),
            zof(mardf_gm1001gr, 'N'), zof(extdf_gm1001gr, 'N')),
    z_A = c(zof(mardf_gm1001g, 'A'), zof(extdf_gm1001g, 'A'),
            zof(mardf_gm1001gr, 'A'), zof(extdf_gm1001gr, 'A'))
)
ztab$z_theory <- zof(theorydf_gm1001gh, 'N')
print(ztab, digits = 3)
```

On the real map, `z_N` and `z_A` disagree with each other and both sit above the
prediction. Once the locations are shuffled, the two columns collapse onto each other and
onto `z_theory`.

# An important take-away: what does the $z$ value capture?

Three separate things influence $z$ in a realistic dataset, and the sections above have
now identified them one by one.

1. **Mutation abundance**, summarized by the SFS. This is the part `MARtheory()`
   predicts, and on `gmexp` it accounts for the entire curve.
2. **Breeding structure.** The SFS counts chromosomes, but we sample individuals, and the
   two coincide only under random mating. In the fully inbred *Arabidopsis* accessions
   the chromosome-based prediction is wrong by tens of percent until the ploidy is
   described correctly.
3. **Spatial structure**, summarized by the $N \sim A$ relationship and by how related
   nearby individuals are. Random placement makes area a clean proxy for sample size and
   forces the area-based and size-based exponents to agree
   (`r sprintf("%.2f", ztab$z_A[3])` and `r sprintf("%.2f", ztab$z_N[3])` on the
   relocated map). Clustered sampling pulls them apart
   (`r sprintf("%.2f", ztab$z_A[1])` versus `r sprintf("%.2f", ztab$z_N[1])` on the real
   map) and lifts both above the SFS prediction of
   `r sprintf("%.2f", ztab$z_theory[1])`.

A $z$ estimated from real data mixes all three together. Reading it as a statement about
genetic diversity requires accounting for the other two first.

# If you are really into the math {#math}

Throughout, let $N$ be the number of individuals in the full dataset, $x$ the ploidy, and
$xN$ the total number of sequenced chromosomes. The genotype matrix covers $L$ sites. At
each site, $k$ denotes the number of chromosomes carrying the mutation, and $\sigma(k)$ is
the proportional site frequency spectrum, so that $\sum_{k=0}^{xN}\sigma(k) = 1$. We
sample $n$ chromosomes, equivalently $j = n/x$ individuals (this $j$ is what the package
reports in the `N` column), and write $H_m = \sum_{i=1}^{m} 1/i$ for the $m$-th harmonic
number.

The goal is the expected number of captured mutations $M$ as a function of $n$, which in
turn stands in for area.

## $n = 1$ and $L = 1$

One chromosome is sequenced at a single site, where the mutation is carried by $k$ of the
$xN$ chromosomes. $M$ can only be 0 or 1, so

$$
M \sim \operatorname{Bernoulli}\!\left(\frac{k}{xN}\right).
$$

## $n = 1$ and $L > 1$

With $L$ sites we have $L$ independent Bernoulli trials, but the success probability
differs from site to site because $k$ does. Treating $k$ as a random variable drawn from
the spectrum, $K \sim \sigma(k)$, the marginal success probability is

$$
\bar{p} \;=\; \sum_{k=0}^{xN} \sigma(k)\,\frac{k}{xN},
$$

and the total number of captured mutations is

$$
M \sim \operatorname{Binom}(L, \bar{p}).
$$

## $n > 1$ and $L > 1$

When more than one chromosome is sampled, the per-site success probability is no longer
$k/xN$; it is one minus a hypergeometric miss probability. Sampling $n$ chromosomes
without replacement from $xN$, the mutation is missed only when all $n$ come from the
$xN-k$ non-carriers:

$$
P(\text{captured}) \;=\; 1 - \frac{\binom{xN-k}{n}}{\binom{xN}{n}}.
$$

$M$ is still binomial, so its expectation is $L$ times the spectrum-averaged capture
probability:

$$
\boxed{\;E[M] \;=\; L \sum_{k=0}^{xN} \sigma(k)
\left(1-\frac{\binom{xN-k}{n}}{\binom{xN}{n}}\right)\;}
$$

## The other two diversity measures

**Watterson's estimator** is defined on sites that are *segregating* in the subsample,
that is $1 \le K \le n-1$. Both fixed outcomes must therefore be removed: the site can
come out monomorphic because no mutation was sampled, or because *only* mutations were
sampled. Dividing the resulting per-site fraction by $H_{n-1}$ gives

$$
E[\theta_w] \;=\; \frac{1}{H_{n-1}} \sum_{k=0}^{xN} \sigma(k)
\left(1-\frac{\binom{xN-k}{n}}{\binom{xN}{n}}-\frac{\binom{k}{n}}{\binom{xN}{n}}\right).
$$

**Nucleotide diversity** needs no hypergeometric projection at all. Because it is an
average over pairs, the spectrum-averaged full-dataset value is already the answer:

$$
E[\theta_\pi] \;=\; \sum_{k=0}^{xN} \sigma(k)\,\frac{2k(xN-k)}{xN(xN-1)}
$$

Note that $n$ does not appear. That is exact rather than approximate, and together with
the $\theta_w$ expression it explains why the two panels above behave so differently.

**Proposition 1. $E[\theta_\pi]$ is invariant in $n$, for any spectrum.** At a single
site, $\hat\theta_\pi = \Delta / \binom{n}{2}$, where $\Delta$ counts discordant pairs
among the sampled chromosomes. By exchangeability of sampling without replacement, each of
the $\binom{n}{2}$ sampled pairs is marginally a uniformly random pair drawn from the
$\binom{xN}{2}$ pairs in the full dataset, so
$E[\Delta] = \binom{n}{2}\,k(xN-k)/\binom{xN}{2}$. The $\binom{n}{2}$ cancels on division,
leaving the full-dataset value for every $n \ge 2$; averaging over sites with weights
$\sigma(k)$ gives the boxed expression. No model assumption enters. The argument is pure
combinatorics, which is why the $\theta_\pi$ curve stays flat regardless of demography or
selection.

**Proposition 2. $E[\theta_w]$ is flat in $n$ if and only if $\sigma(k) \propto 1/k$.**
The neutral spectrum is a fixed point of hypergeometric projection. Using
$\frac{1}{k}\binom{k}{i} = \frac{1}{i}\binom{k-1}{i-1}$ and the Vandermonde convolution
$\sum_m \binom{m}{a}\binom{M-m}{b} = \binom{M+1}{a+b+1}$, a spectrum $\phi_k = \theta/k$
projects onto a subsample of size $n$ as $\tilde\phi_i = \theta/i$ for $i = 1,\dots,n-1$.
Summing gives $E[S_n] = \theta H_{n-1}$, so dividing by $H_{n-1}$ returns $\theta$ at
every $n$. For any other spectrum the cancellation is incomplete. In general
$\theta_w(n)$ runs from $\theta_\pi$ at $n = 2$ to the full-dataset $\theta_w$ at
$n = xN$, so the direction of its trend is the sign of $-D$, where $D$ is Tajima's $D$.
An upward drift, as in `gmexp`, means the dataset carries an excess of rare variants.

### When random mating fails

Everything above samples chromosomes. Under random mating that is equivalent to sampling
individuals, because an individual is $x$ independent gene copies. Under inbreeding it is
not, and the resulting error can be written down exactly. For a fully inbred, completely
homozygous, diploid panel ($F = 1$), sampling $j$ individuals gives

$$
\frac{E[\hat\theta_{\pi,j}]}{\theta_{\pi,\text{full}}}
\;=\; \frac{1 - \dfrac{1}{2j-1}}{1 - \dfrac{1}{2N-1}},
$$

which is again independent of the spectrum. The reason is that the $2j$ sampled
chromosomes contain $j$ within-individual pairs that are always identical and can never
contribute a difference. Those pairs make up a fraction $1/(2j-1)$ of all
$\binom{2j}{2}$ pairs, against $1/(2N-1)$ in the full panel, and the bias is exactly the
difference between the two fractions. The approach to the true value goes as $1/(2j)$:
with $N = `r length(gm1001g$maps$sample.id)`$, a 33% deficit at $j = 2$ and still 0.08% at
$j = 400$. This is the same correlation between the two copies within an individual that
made the diploid prediction overshoot in the *Arabidopsis* section, and recasting the data
as haploid is what removes it.

## So which $z$ value am I comparing?

Putting the pieces together: the scaling of $M$ with $n$ is determined by four
quantities, the number of sites $L$, the number of chromosomes $xN$, the spectrum
$\sigma(k)$, and the sample size $n$. For real data, two further factors sit between that
prediction and the observed mutations-area curve: the breeding structure, which controls
whether sampling individuals is the same as sampling chromosomes, and the spatial
structure, which controls whether sampling area is the same as sampling individuals. A
$z$ reported without saying which of the three it is meant to reflect is not comparable
across datasets, and `MARtheory()` is the tool for separating the first from the other
two.
