---
title: "The sampling theory behind MAR"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{The sampling theory behind MAR}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown:
        wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    fig.align = "center",
    out.width = "100%"
)
rm(list = ls())
devtools::load_all()
```

# Why does the mutations-area relationship exist?

In macroecology, species-area relationships (SAR) can be described exhaustively from species abundance distributions (SAD) and species spatial distribution (Tjørve et al. 2008), and extensive studies have explored the theoretical relationship between the species abundance distribution and species-area relationship over the past century (Preston 1948, 1962; May 1975; Harte, Kinzig, and Green 1999; Hubbell 2001). Extending this framework in population genetics, we demonstrate that a similar sampling relationship exists between the site frequency spectrum (SFS), a fundamental population genetics summary statistics, and the mutations-area relationship (MAR). 

In this vignette, we first establish the theoretical connection between the SFS and the SAD as scaled probability density functions of species or mutation abundances, and note one small distinction. 
Then, we introduce the `MARtheory()` function. Assuming that individuals are **randomly mating and randomly distributed** in the population, `MARtheory()` derive the exact mutations-area relationship using only the site frequency spectrum, and the population density $p$, which relates total number of sampled individuals is proportional to area: N = pA (Preston 1962). Without need to know about each individual's specific location, or to run the `MARsampling()` procedures, knowing the SFS, we will have the MAR. `MARtheory()`'s theoretical prediction is then validated using `MARsampling()` on a simulated dataset (`gmexp` data).

Next, we show that when the two key assumptions of **random mating and random distributions*8 are violated in real life, in our *Arabidopsis thaliana* 1001G example (`gm1001g` data), how the theory prediction deviates from the `MARsampling()` output. And when we treat the 1001g data as haploid individuals and shuffle the sample locations, `MARtheory()` predicts `MARsampling()` output in a real-life dataset.

In the end, by show that the MAR can be theoretically derived using the SFS, we discuss its profound implication on interpreting the $z$ values for conservation.

## Some assumptions and the example data

We start by simulating 50 diploid individuals sampled from a randomly mating and randomly distributed population that is growing exponentially using the python package `msprime` (code available in `data-raw/gmexp_simulate.py`). We constructed a `genomaps` object for this simulated population (code available in `data-raw/gmexp_rda.R`), and provided it with this package.

You can explore the data here:

```{r explore gmexp}
data(gmexp)
print(gmexp)
```

To elaborate on the two key assumptions for deriving the MAR from SFS: 

**1. Individuals are randomly mating.** Under random mating an individual is just
`ploidy` independently drawn gene copies, so *sampling one diploid individual is the same
as sampling two random chromosomes*. This equivalence allowed us to use the SFS, which is bookkeeping at chromosome level, while actually collecting individuals. 

**2. Individuals are randomly distributed in space.** This is Preston's classical
assumption, and it is what allows area to stand in for sample size: if individuals are
scattered at random, a box twice as large holds twice as many of them, so $N \propto A$.
The `gmexp` coordinates were drawn from a uniform distribution, which we can see by plotting the `marmaps` object stored within the `genomaps` object.

```{r randomloc}
plot(gmexp$maps)
```

Both assumptions are essential for the derivation later. 
In the *Arabidopsis* section we violate each assumption in turn and
measure what it costs.

## What is a site-frequency spectrum?

Before we dive into the sampling relationship, we need to explain what is a SFS. 

Take a genetic dataset with $N$ individuals of ploidy $x$, so there are $xN$
chromosomes in total, genotyped at $L$ sites. At any one genomic site, count how many of the
$xN$ chromosomes carry the mutation and call that count $k$. The **site-frequency
spectrum** (SFS) is the tally of how many sites have $k = 1$ (singletons), how many have $k = 2$, and
so on. We write the proportional site frequency spectrum as $\sigma(k)$.

For `gmexp` that is $L = `r length(gmexp$geno$allele_count)`$ sites spread over
$xN = `r length(gmexp$maps$sample.id) * gmexp$geno$ploidy`$ chromosomes:

```{r sfsplot}
obs <- sfs(gmexp)
plot(obs)
```

The mutations-area relationship arises from a core principle: the commonness of rarity. Most mutations in a population are rare, and only a few reach high frequencies.

The site frequency spectrum (SFS) captures this distribution of allele frequencies and is a fundamental summary statistic in population genetics. It represents the number of mutations observed at different frequencies within a sample. The shape of the SFS, which often skewed toward low-frequency variants, is what gives rise to the scaling behavior observed in MAR.

This relationship mirrors a well-known pattern in ecology: just as the species-area relationship (SAR) emerges from the species abundance distribution (SAD), the mutations-area relationship (MAR) is a consequence of the site frequency spectrum (SFS), both rooted in the statistical structure of rarity in biological systems.

However, **a key distinction between the
SFS and the SAD** is that the sum of SAD entries is the total number of sampled
individuals, whereas the sum of SFS entries is the total number of mutations, which is
unrelated to the number of individuals sampled Any species-abundance model can be fitted
to an SFS, since both are probability densities over abundance classes, but they are not
the same object. For a direct comparison to a SAD, a haplotype abundance distribution
under the infinite-alleles assumption would be more appropriate.

Excitingly, the mutations-area relationshp can emerge from any shape of the spectrum as long as the individuals are randomly mating and randomly distributed. The individuals may be under
selection or have passed through demographic change; we simply take the spectrum as
given. `gmexp` illustrates this well precisely because it is *not* a population evolving under : it grew
exponentially, and exponential growth leaves an excess of rare variants. We can check
that against the neutral expectation the package ships, `expsfs()`:

```{r neutral}
plot(obs, expected = gmexp)
# pass the `log = 'x'` option to inspect the singletons
plot(obs, expected = gmexp, log = 'x')
# pass the `log = 'y'` option to inspect the mutation abundance (graphics::plot() will warns about zero-entries)
plot(obs, expected = gmexp, log = 'y')
```

Singletons are more common than the
constant-size neutral model predicts, and high-frequency variants are depleted. That is the
exponential growth showing up in the spectrum, and it does not interfere with the
sampling theory as we will show below.

## How to build the MAR from a site-frequency spectrum?

### `MARtheory()` implements an analytical sampling process result 

Recall the two assumptions: individuals mate at random, so sampling $n$ chromosomes is
the same as sampling $n/x$ individuals; and individuals distribute raomdomly in space, so
sampling a larger area leads to sampling more individuals.

Consider a single genomic site at which $k$ of the $xN$ chromosomes carry the mutation $m_i$, and
suppose we draw $n$ chromosomes. We miss to sample the mutation $m_i$ only if every one of the $n$ draws
came from the $xN - k$ chromosomes that do not carry it. Counting the ways that can
happen gives

$$
P(\text{missed}) \;=\; \frac{\binom{xN - k}{n}}{\binom{xN}{n}},
$$

so we sample the site with probability $1 - P(\text{missed})$. An mutation carried by
half the population ($k = xN/2$) is more likely to be sampled, whereas a singleton mutation ($k = 1$) is rarely sampled, and more likely to be missed. Averaging this sampling probability across multiple sites with varying mutation frequency gives the expected number of mutations in a sample of $n$ chromosomes. The exact derivations are available in [the math section](#math).

We can therefore predict the whole mutations-area relationship curve using the equation below, which is implemented in `MARtheory()`. 

$$
\boxed{\;E[M] \;=\; L \sum_{k=0}^{xN} \sigma(k)
\left(1-\frac{\binom{xN-k}{n}}{\binom{xN}{n}}\right)\;}
$$



`MARtheory()` computes the above sampling relationship purely from theory without running any sampling scheme, and we can fit the mutations-area relationship using `MARcalc()`. 

```{r marexp}
expdf <- MARtheory(gmexp)
head(expdf)

# plot the expdf
plot(expdf)

# run MARcalc()
expdf_fit <- MARcalc(expdf, Mtype = 'M',Atype = 'N')
summary(expdf_fit)

# overlay the fitted MARcalc results on the plot
plot(expdf, fit = expdf_fit$par)

# you don't have to run MARcalc() separately, supply `fit = TRUE`, and the plot method for `martheory` data class will automatically fit and overlay the line. 
plot(expdf, fit = TRUE)
```

### `MARsampling()` confirms `MARtheory()` prediction
 
 we can compare it against the sampling output, which builds the same curve the hard way by repeatedly sampling boxes on the map:

```{r marexp compare}
mardf <- MARsampling(gmexp, myseed = 1)
head(mardf)
plot(mardf, Mtype = "M", Atype = "N", theory = expdf)
```

The blue prediction lands on top of the sampled points, and to either `MARsampling()` or `MARtheory()` output. The scaling factor $z$ from `MARsampling()` is very close to the scaling factor from `MARtheory()`. 

```{r zexp}
# fit the MAR
mardf_fit <- MARcalc(mardf, Mtype = 'M',Atype = 'N')

summary(mardf_fit)

# extract the fitted c and z value
print(mardf_fit$par)
print(expdf_fit$par)

# you can compare the theory prediction with the empirical fit side-by-side
plot(mardf, Mtype = "M", Atype = "N", theory = expdf, fit = mardf_fit$par)

# or use a log-log scale
plot(mardf, Mtype = "M", Atype = "N", theory = expdf, fit = mardf_fit$par, logscale = TRUE)
```

### Predictions for Watterson's $\theta$ and nucleotide diversity

Using the same sampling mechanism, we can predict the Watterson's $\theta$ ($\theta_w$) and nucleotide diversity ($\theta_\pi$) measures as well. 

Watterson's $\theta_w$ increases with the sample size. It counts
segregating sites and divides by a correction factor which accounts for the sample size. 

$$
Fill in theta_w
$$

Under neutral coalescent theory, $\theta_w$ is an unbiased estimator for population scaled mutation rate. However, since our simulated `gmexp` object is based on an exponentiallly growing population, $\ that is only the right
correction when the spectrum has the neutral $1/k$ shape. This spectrum does not, so the
correction fails to cancel and the estimate changes with sample size.

```{r theta_w}
plot(mardf, Mtype = "thetaw", Atype = "N", theory = expdf)
```

Nucleotide diversity $\theta_\pi$ does not change with sample size, taking exactly. It averages over *pairs* of chromosomes, and any pair we draw is just a
random pair from the full dataset, so subsampling cannot shift it. 

```{r thetapi}
plot(mardf, Mtype = "thetapi", Atype = "N", theory = expdf)
```

### Going from sampling individuals to sampling areas 

Because the individuals are randomly placed, sample size and area really are
interchangeable here:

```{r nagmexp}
plot(A ~ N, data = mardf)
N_A_lm <- lm(A ~ N, data = mardf)
summary(N_A_lm)
# extract the scaling factor
p <- N_A_lm$coefficients['N']
```

Therefore, we can generate the expected MAR and compare it with sampled MAR. We did not include this last step of going from N to A as this would require fitting of the `MARsampling()` results. But we note that this is still purely analytical, as we did not use any sampled mutation / genetic diversity values and only used the population density information. 

```{r add area}
# Add the expected area 
expdf$A = expdf$N * p

# Overlay with the MARsampling output 
plot(mardf)
lines(M ~ A, data = expdf, col = 'blue')
```


# How does this work in our arabidopsis dataset?

The *Arabidopsis thaliana* 1001 Genomes accessions break both assumptions at once, which
makes them a good case for seeing how the scaling factor $z$ is determined by .

```{r loadara}
data(gm1001g)
ara <- gm1001g
c(individuals = length(ara$maps$sample.id), sites = length(ara$geno$allele_count))
```

Run the prediction and compare it against random subsamples of individuals. We count
mutations the same way `MARsampling()` does, scoring a site as captured once at least one
sampled individual carries the mutation:

```{r arabroken}
count_M <- function(gm, idx) {
    sum(matrixStats::rowSums2(gm$geno$genotype, cols = idx) > 0)
}

set.seed(2)
js <- c(1, 5, 25, 100, 500)
sampled <- sapply(js, function(j) {
    mean(replicate(500, count_M(ara, sample.int(length(ara$maps$sample.id), j))))
})

arath <- MARtheory(ara)
data.frame(N = js, sampled = sampled, predicted = arath$M[js],
           pct_error = 100 * (arath$M[js] / sampled - 1))
```

The prediction overshoots by roughly half at the smallest sample sizes, and the gap only
closes at the largest ones.

## It acts as a haploid

The genotypes explain why. These accessions are inbred lines, and they are *completely
homozygous*:

```{r homozygous}
mean(ara$geno$genotype == 1) # fraction of heterozygous calls
```

Not one heterozygote. The same fact shows up in the spectrum as a striking regularity:
every odd allele count is empty, because a mutation can only ever be carried in pairs.

```{r oddk}
arasfs <- as.vector(sfs(ara, nozero = FALSE))
sum(arasfs[seq(2, length(arasfs) - 1, by = 2)]) # total count at odd k
```

This breaks assumption 1 outright. The two chromosomes inside an accession are not two
independent draws, they are identical copies. When we sample one individual and hand the
theory two chromosomes, it believes it had two independent chances to catch each mutation
when it really had one, so it predicts too many.

The remedy is to describe the data as what it effectively is: a haploid sample of
`r length(ara$maps$sample.id)` gene copies rather than a diploid sample of
`r 2 * length(ara$maps$sample.id)`.

```{r haploid}
arah <- ara
arah$geno$genotype <- ara$geno$genotype / 2
arah$geno$ploidy <- 1
arah$geno$allele_count <- matrixStats::rowSums2(arah$geno$genotype, na.rm = TRUE)

haplth <- MARtheory(arah)
data.frame(N = js, sampled = sampled, predicted = haplth$M[js],
           pct_error = 100 * (haplth$M[js] / sampled - 1))
```

Tens of percent become a fraction of one percent. Assumption 1 accounted for all of it.

Note that the package has no function for this recast, so it has to be done by editing
the `margeno` list directly. `margeno()` validation would reject the rescaled matrix on
the way in.

That repairs the relationship between mutations and *sample size*. The MAR is about
*area*, though, and for these accessions the two are no longer interchangeable:

```{r naara, fig.width = 5, fig.height = 4}
aradf <- MARsampling(ara, myseed = 1, nrep = 5)
plot(N ~ A, data = aradf)
fita <- lm(log(N) ~ log(A), data = aradf[!is.na(aradf$N) & aradf$N > 0, ])
coef(fita)[2]
```

The slope is `r sprintf("%.3f", coef(fita)[2])`, not 1. Accessions are clustered, with
some regions sampled heavily and others barely, so doubling the area does not double the
number of accessions inside it. Assumption 2 is broken too.

## Now we move it around so it appears randomly located

To isolate the spatial effect, hold the genotypes fixed and give every accession a new,
uniformly random location:

```{r relocate}
set.seed(42)
nind <- length(ara$maps$sample.id)
randloc <- data.frame(id = ara$maps$sample.id,
                      lon = runif(nind, 0, 1000),
                      lat = runif(nind, 0, 1000))
mapsrand <- marmaps(lonlatdf = randloc, mapcrs = "", incrs = "")
ararand <- genomaps(geno = arah$geno, maps = mapsrand)

randdf <- MARsampling(ararand, myseed = 1, nrep = 5, quorum = FALSE)
```

```{r narand, fig.width = 5, fig.height = 4}
plot(N ~ A, data = randdf)
fitr <- lm(log(N) ~ log(A), data = randdf[!is.na(randdf$N) & randdf$N > 0, ])
coef(fitr)[2]
```

The slope snaps back to `r sprintf("%.3f", coef(fitr)[2])`. Now that area and sample size
carry the same information, the exponent stops depending on which of them we regress
against, and the haploid theory recovers it:

```{r zcompare}
zz <- function(df, Atype) unname(MARcalc(df, Mtype = "M", Atype = Atype)$par["z"])
data.frame(
    dataset = c("Arabidopsis, real locations", "Arabidopsis, relocated at random"),
    z_area  = c(zz(aradf, "A"), zz(randdf, "A")),
    z_N     = c(zz(aradf, "N"), zz(randdf, "N"))
)

zz(haplth, "N") # haploid theory prediction
```

On the real, clustered map the two exponents disagree. Once the accessions are randomly
located they agree with each other and with the prediction made from the spectrum alone.

One caveat on units: `A` sits on an arbitrary scale here because these objects carry no
CRS, so `z` is comparable across the objects but the intercept `c` is not.

# An important take-away: what does the z value capture?

Three separate things feed into `z` in a realistic dataset, and the sections above have
now pulled them apart.

1. **Genotype abundance**, summarized by the SFS. This is the part `MARtheory()`
   predicts, and on `gmexp` it accounts for the entire curve.
2. **Breeding structure.** The SFS counts chromosomes, but we sample individuals, and the
   two coincide only under random mating. In the fully inbred *Arabidopsis* accessions
   the chromosome-based prediction is wrong by tens of percent until the ploidy is
   described correctly.
3. **Spatial structure**, summarized by the $N \sim A$ relationship. Random placement
   gives a slope of 1 and makes area a clean proxy for sample size. Clustered sampling
   (slope `r sprintf("%.2f", coef(fita)[2])` here) means the area-based and size-based
   exponents are measuring different things.

A `z` estimated from real data mixes all three together. Reading it as a statement about
genetic diversity requires accounting for the other two first.

## If you are really into the math {#math}

Throughout, let $N$ be the number of individuals, $x$ the ploidy, and $xN = cN$ the total
number of sequenced chromosomes. The genotype matrix covers $L$ sites. At each site, $k$
denotes the number of chromosomes carrying the mutation, and $\sigma(k)$ is the
proportional site-frequency spectrum, so that $\sum_{k=0}^{xN}\sigma(k) = 1$. We sample
$n$ chromosomes and write $H_m = \sum_{i=1}^{m} 1/i$ for the harmonic number.

The goal is the expected number of captured mutations $M$ as a function of $n$, which in
turn stands in for area.

### $n = 1$ and $L = 1$

One chromosome is sequenced at a single site, where the mutation is carried by $k$ of the
$xN$ chromosomes. $M$ can only be 0 or 1, so

$$
M \sim \operatorname{Bernoulli}\!\left(\frac{k}{xN}\right).
$$

### $n = 1$ and $L > 1$

With $L$ sites we have $L$ independent Bernoulli trials, but the success probability
differs from site to site because $k$ does. Treating $k$ as a random variable drawn from
the spectrum, $K \sim \sigma(k)$, the marginal success probability is

$$
\bar{p} \;=\; \sum_{k=0}^{xN} \sigma(k)\,\frac{k}{xN},
$$

and the total number of captured mutations is

$$
M \sim \operatorname{Binom}(L, \bar{p}).
$$

### $n > 1$ and $L > 1$

When more than one chromosome is sampled, the per-site success probability is no longer
$k/xN$; it is one minus a hypergeometric miss probability. Sampling $n$ chromosomes
without replacement from $xN$, the mutation is missed only when all $n$ come from the
$xN-k$ non-carriers:

$$
P(\text{captured}) \;=\; 1 - \frac{\binom{xN-k}{n}}{\binom{xN}{n}}.
$$

$M$ is still binomial, so its expectation is $L$ times the spectrum-averaged capture
probability:

$$
\boxed{\;E[M] \;=\; L \sum_{k=0}^{xN} \sigma(k)
\left(1-\frac{\binom{xN-k}{n}}{\binom{xN}{n}}\right)\;}
$$

### The other two diversity measures

**Watterson's estimator** is defined on sites that are *segregating* in the subsample,
that is $1 \le K \le n-1$. Both fixed outcomes must therefore be removed: the site can
come out monomorphic because no mutation was sampled, or because *only* mutations were
sampled. Dropping the second term is a subtle error, because it leaves $\theta_w$
non-flat even for a perfectly neutral spectrum, where it must equal $\theta$ by
construction.

$$
E[\theta_w] \;=\; \frac{1}{H_{n-1}} \sum_{k=0}^{xN} \sigma(k)
\left(1-\frac{\binom{xN-k}{n}}{\binom{xN}{n}}-\frac{\binom{k}{n}}{\binom{xN}{n}}\right)
$$

**Nucleotide diversity** needs no hypergeometric projection at all:

$$
E[\theta_\pi] \;=\; \sum_{k=0}^{xN} \sigma(k)\,\frac{2k(xN-k)}{xN(xN-1)}
$$

Note that $n$ does not appear. That is exact rather than approximate, and it leads to the
two results that explain why the two panels above behave so differently.

**Proposition 1. $E[\theta_\pi]$ is invariant in $n$, for any spectrum.**
$\theta_\pi$ is an average over pairs: $\hat\theta_\pi = D / \binom{n}{2}$, where $D$
counts discordant pairs among the sampled chromosomes. By exchangeability of sampling
without replacement, each of the $\binom{n}{2}$ sampled pairs is marginally a uniformly
random pair drawn from the $\binom{xN}{2}$ pairs in the full dataset, so
$E[D] = \binom{n}{2}\,k(xN-k)/\binom{xN}{2}$. The $\binom{n}{2}$ cancels on division,
leaving the full-dataset value for every $n \ge 2$. No model assumption enters. The
argument is pure combinatorics, which is why the $\theta_\pi$ curve stays flat regardless
of demography or selection.

**Proposition 2. $E[\theta_w]$ is flat if and only if $\sigma(k) \propto 1/k$.**
The neutral spectrum is a fixed point of hypergeometric projection. Using
$\frac{1}{k}\binom{k}{j} = \frac{1}{j}\binom{k-1}{j-1}$ and the Vandermonde convolution
$\sum_m \binom{m}{a}\binom{M-m}{b} = \binom{M+1}{a+b+1}$, a spectrum $\phi_k = \theta/k$
projects onto a subsample of size $n$ as $\tilde\phi_j = \theta/j$ for $j = 1,\dots,n-1$.
Summing gives $E[S_n] = \theta H_{n-1}$, so dividing by $H_{n-1}$ returns $\theta$ at
every $n$. For any other spectrum the cancellation is incomplete. In general
$\theta_w(n)$ runs from $\theta_\pi$ at $n = 2$ to the full-dataset $\theta_w$ at
$n = xN$, so the direction of its trend is the sign of $-D$, where $D$ is Tajima's $D$.
An upward drift, as in `gmexp`, means the dataset carries an excess of rare variants.

### When random mating fails

Everything above samples chromosomes. Under random mating that is equivalent to sampling
individuals, because an individual is $x$ independent gene copies. Under inbreeding it is
not, and the resulting error can be written down exactly. For a fully inbred, completely
homozygous panel ($F = 1$), sampling $j$ individuals gives

$$
\frac{E[\hat\theta_{\pi,j}]}{\theta_{\pi,\text{full}}}
\;=\; \frac{1 - \dfrac{1}{2j-1}}{1 - \dfrac{1}{2N-1}},
$$

which is again independent of the spectrum. The reason is that the $2j$ sampled
chromosomes contain $j$ within-individual pairs that are always identical and can never
contribute a difference. Those pairs make up a fraction $1/(2j-1)$ of all pairs, against
$1/(2N-1)$ in the full panel, and the bias is exactly the difference between the two
fractions. The approach to the true value goes as $1/(2j)$: a 33% deficit at $j = 2$, and
still 0.08% at $j = 400$. This is the same correlation between the two copies within an
individual that made the diploid prediction overshoot by half in the *Arabidopsis*
section, and recasting the data as haploid is what removes it.

### Summary

The scaling of $M$ with $n$ is determined by four quantities: the number of sites $L$,
the number of chromosomes $xN$, the spectrum $\sigma(k)$, and the sample size $n$. For
real data, two further factors sit between that prediction and the observed
mutations-area curve: the breeding structure, which controls whether sampling individuals
is the same as sampling chromosomes, and the spatial structure, which controls whether
sampling area is the same as sampling individuals.
